---
title: "LGEO2185: Session 4 -- Machine Learning in Geosciences"
author: "Kristof Van Oost & Antoine Stevens"
Date: 21/02/2026
editor: 
  markdown: 
    wrap: 72
---

::: callout-note
This session assumes proficiency with geographic data analysis, basic R,
for example gained by studying the contents and working through the
previous session. A familiarity with geospatial approaches ad spatial
modelling is recommended.
:::

# Introduction

## What Do We Mean by “Spatial Models”?

Spatial modelling spans a continuum: from models that use spatial data
to models that explicitly represent spatial dependence, interaction, or
physical processes.

-   **Statistical models that use geographical data (space as data)**
    -   Each grid cell treated as independent
    -   Examples: GLM, GAM, Random Forest, logistic regression on
        rasters, ML on spatial data (see further)
    -   Use spatial predictors, but ignore spatial autocorrelation
-   **GIS-based models (space as operations)**
    -   Chain spatial tools and map algebra operations
    -   Examples: weighted overlay, buffering, rule-based suitability
        models
    -   No explicit modelling of dependence or uncertainty
-   **Spatial statistics (space as dependence)**
    -   Explicitly model spatial autocorrelation
    -   Examples: spatial lag/error models, CAR/SAR, spatial GLMM
-   **Geostatistics (space as continuous field)**
    -   Assume a spatial covariance structure
    -   Examples: variogram, kriging, Gaussian processes
-   **Spatial AI (space as learned pattern)**
    -   Learn spatial structure implicitly from pixel neighborhoods
    -   Examples: CNNs, U-Net, vision transformers
-   **Agent-based models (space as interaction environment)**
    -   Agents interact locally in space
    -   Emergent spatial patterns arise from micro-level decisions
    -   Examples: land-use change, socio-ecological systems
-   **Process-based models (space as physical system)**
    -   Mechanistic simulation of spatial processes
    -   Examples: erosion, hydrology, landscape evolution models

------------------------------------------------------------------------

::: callout-note
A model becomes truly spatial when it: Models spatial dependence
explicitly; Simulates spatial interactions; Represents spatial processes
mechanistically
:::

## What is Machine Learning (and how does it relate to AI)?

Artificial Intelligence (AI) refers to technology that simulates (or in
some cases exceeds) human behaviour, including:

-   Learning
-   Comprehension
-   Problem solving
-   Decision making
-   Autonomy
-   Creativity

A classical concept in AI is the **Turing Test**, which states that:

> A machine is intelligent if a human cannot reliably distinguish its
> behaviour from that of a human.

A more modern, data-driven definition (and the one adopted in this
course) is:

> **Artificial Intelligence is the design of algorithms that enable
> machines to learn patterns from data and use them to perform cognitive
> tasks such as classification, prediction, and reasoning.**

------------------------------------------------------------------------

Machine learning (ML) is a subset of AI focused on creating models by
training algorithms to make predictions or decisions based on data. In
essence, it can be seen as **advanced statistical learning from past
observations**. ::: callout-note Machine learning enables a machine to
automatically learn from data, improve performance from experiences, and
predict things without being explicitly programmed :::

Machine learning encompasses a broad range of techniques that allow
computers to learn from data **without being explicitly programmed**.
Common examples include:

-   Linear regression
-   Logistic regression
-   Decision trees
-   Random forest
-   Support vector machines
-   k-nearest neighbours
-   Clustering methods

Machine learning is commonly divided into:

-   **Supervised learning**, where training data are labelled by humans\
-   **Unsupervised learning**, where patterns or labels are inferred
    automatically

The core objective of machine learning is simple: to learn the mapping
between inputs and outputs using training data, so that labels or values
can be predicted for new, unseen data.

A useful overview of the relationship between AI, machine learning, deep
learning, and generative AI is available
[here](https://assets.ibm.com/is/image/ibm/diagram-comparing-ai-ml-deep-learning-gen-ai:16x9?fmt=png-alpha&dpr=on%2C2&wid=1536&hei=864):

------------------------------------------------------------------------

## And Deep Learning?

Deep learning is a subset of machine learning that uses **multi-layered
neural networks** to automatically learn hierarchical representations of
data.

By stacking many processing layers, deep learning models can capture
complex, non-linear relationships and are particularly effective for
images, time series, and text.

------------------------------------------------------------------------

::: callout-note
If you want to know more about machine learning and deep learning
approaches in geosciences (including hybrid approaches and integration
with physically based models), here is a good review paper: Reichstein,
M., Camps-Valls, G., Stevens, B. et al. Deep learning and process
understanding for data-driven Earth system science. Nature 566, 195–204
(2019). <https://doi.org/10.1038/s41586-019-0912-1>
:::

## How does Machine Learning work?

Machine learning model development is an iterative process and is rarely
a linear process. Instead, it follows an iterative cycle of exploration,
refinement, comparison, and reassessment. The figure below illustrates a
typical pathway toward identifying an appropriate model. The main phases
are.

-   **Exploratory Data Analysis (EDA)** Model development begins with
    exploration. Numerical summaries and visualization interact in a
    back-and-forth process where early discoveries generate new
    questions and additional analyses.
    -   Identify patterns and relationships
    -   Detect anomalies or outliers
    -   Assess distributions and transformations
    -   Reveal potential modeling challenges.
-   **Feature Engineering** Insights from EDA guide the construction of
    meaningful predictors that improve model representation.
    -   Transform variables
    -   Combine predictors (e.g., ratios or indices)
    -   Apply dimensionality reduction (e.g., PCA)
    -   Derive spatial or temporal aggregates Effective feature
        engineering often matters more than algorithm choice.
-   **Model Tuning and Selection** Multiple candidate models are built
    and compared. Some require tuning of structural hyperparameters.
    Performance is assessed through resampling strategies such as
    cross-validation, ensuring robust comparison.
    -   Fit model
    -   Tune parameters
    -   Re-evaluate
    -   Compare alternatives This stage is inherently iterative.
-   **Model Evaluation** Promising models are critically assessed using
    quantitative and diagnostic tools.
    -   Examine performance metrics (RMSE, AUC, R²)
    -   Inspect residual diagnostics
    -   Conduct additional exploratory analyses
    -   Perform formal between-model comparisons Differences between
        models must be interpreted relative to experimental uncertainty.

Model development is cyclical. Exploration, feature construction,
tuning, and evaluation continuously inform one another until a
defensible and well-understood model is obtained.

![Model development and selection workflow. Concept adapted from: Kuhn &
Silge (2022), *Tidy Modeling with R*.
https://www.tmwr.org/software-modeling](images/ML_pipeline.png){fig-align="center"
width="80%"}

## 1. Why do we evaluate/validate models?

This session focuses on supervised ML techniques in which there is a
training dataset, as opposed to unsupervised techniques such as
clustering. Response variables can be binary (such as landslide or
species occurrence), categorical (land use), integer (species richness
count) or numeric (soil acidity measured in pH). Supervised techniques
model the relationship between such responses and one or more
predictors. When we fit a model, a key question is:

> How well will it predict *new, unseen data*?

A model can look excellent on the same data used for training
(in-sample), yet perform poorly on new data (overfitting, leakage, or
mismatch between training and prediction conditions). **Validation** is
therefore about *generalization*, not goodness-of-fit.

### Common validation strategies

-   **Apparent performance** (“train = test”): this might give an
    optimistic baseline.
-   **Hold-out split** (e.g., 70/30): a simple approach but potentially
    noisy (depends on the split).
-   **k-fold cross-validation (CV)**: more stable average performance.
-   **LOOCV** (Leave-One-Out CV): extreme form of CV; computationally
    heavier.
-   **Spatial CV**: required when there is spatial dependence; random CV
    can overestimate performance for geographic data.

In this session, we will demonstrate these concepts using a case-study.

Primary inputs (for concepts and best practices):

-   Geocomputation with R — spatial CV chapter:
    https://r.geocompx.org/spatial-cv
-   Sun et al. (2023) Spatial cross-validation for GeoAI (PDF):
    https://www.acsu.buffalo.edu/\~yhu42/papers/2023_GeoAIHandbook_SpatialCV.pdf
-   Rebecca Barter (tuning logic / leakage):
    https://rebeccabarter.com/blog/2020-03-25_machine_learning#tune-the-parameters

------------------------------------------------------------------------

## 2. Setup tutorial data

The data used in this tutorial is derived from Lovelace et al.:
Conceptual explanations in this document are inspired by Geocomputation
with R (Lovelace et al.), (https://r.geocompx.org/). Install the
following packages:

```{r, echo = FALSE, eval = FALSE}
pkgs <- c(
  "sf","terra","dplyr","pROC","spDataLarge",
  "effects","ggplot2",
  "blockCV","spdep"
)

to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install)

install.packages("spDataLarge", repos = "https://geocompr.r-universe.dev")
```

```{r}
library(sf)
library(terra)
library(dplyr)
library(pROC)
library(spDataLarge)
library(effects)
library(ggplot2)
```

------------------------------------------------------------------------

## 3. Data and first visual exploration

We use the `spDataLarge` landslide case study (points + raster
covariates), frequently used in spatial CV teaching materials. The data
is described in detail in @muenchow2012.

```{r}
data("lsl", "study_mask", package = "spDataLarge")
ta <- terra::rast(system.file("raster/ta.tif", package = "spDataLarge"))

head(lsl)
```

The above code loads a dataframe with the observation point data, a `sf`
object named `study_mask` and a `SpatRaster` called `ta` that contains
terrain attribute rasters. Explore this data, you will see that `lsl`
contains the coordinates of observed landslides (ie `TRUE` in the
`lslpts` columns represents a landslide initiation point). The `FALSE`
observations represent non-landslide points that were sampled randomly
from the study area. Use `summary(lsl\$lslpts)` to explore the dataset.
Note that `lslpts` is a factor (ie `0` or `1`).

```{r, echo = FALSE}
# 1) elevation layer (check layer name in your raster stack)
elev_rast <- ta[["elev"]]
elev_masked <- mask(elev_rast, study_mask)

# 2) hillshade
slope_r <- terrain(elev_masked, v = "slope", unit = "radians")
aspect_r <- terrain(elev_masked, v = "aspect", unit = "radians")
hill <- shade(slope_r, aspect_r, angle = 45, direction = 315)

# 3) to df for ggplot
hill_df <- as.data.frame(hill, xy = TRUE)
colnames(hill_df) <- c("x", "y", "hillshade")

# 4) points to sf
lsl_sf <- st_as_sf(lsl, coords = c("x", "y"), crs = st_crs(study_mask))

# 5) plot
ggplot() +
  geom_raster(data = hill_df, aes(x = x, y = y, fill = hillshade)) +
  scale_fill_gradient(low = "black", high = "white") +
  geom_sf(data = study_mask, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = lsl_sf, aes(color = lslpts), size = 2) +
  scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "red")) +
  coord_sf() +
  theme_minimal() +
  labs(title = "Landslide points over hillshade", fill = "Hillshade", color = "Landslide")
```

------------------------------------------------------------------------

## 4. Baseline model: logistic regression (GLM)

We model a binary outcome `lslpts` with a binomial GLM. A Generalized
Linear Model (GLM) extends ordinary linear regression to situations
where the response variable is not continuous and normally distributed
(see session 3). In our case:

-   lslpts is binary
    -   1 = landslide occurred
    -   0 = no landslide
-   We therefore assume a binomial distribution for the response.

We use a logit link function, leading to logistic regression. Binomial
GLMs are widely used in geomorphology and hazard mapping because they:

-   naturally handle presence/absence data
-   produce probabilistic outputs (susceptibility)
-   are interpretable (unlike many black-box ML models)
-   integrate cleanly with spatial prediction on rasters

You can find more information on logistic regression in landslide
susceptibility here @AyalewYamagishi2005. Here, they also serve as a
baseline for comparison with ML methods (see further). To model the
landslide susceptibility, we need predictors. Usually terrain attributes
are good predictors of landsliding. The 'lsl' dataframe already has the
key terrain attributes for all the observations ready for you. These
are:

-   `slope`: slope angle (°)
-   `cplan`: plan curvature (rad m^-1^)
-   `cprof`: profile curvature (rad m^-1^)
-   `elev`: elevation above sea level (m.a.s.l.)
-   `log10_carea`: log10 of the catchment area (log10m^2^).

Note that you could extract these topographic attributes with the help
of R-GIS bridges to SAGA, QGIS etc.

In 'R', model functions can be specified, we use
`lslpts ~ slope + cplan + cprof + elev + log10_carea` and then fit the
logistic regression. It is worth understanding each of the three input
arguments:

-   A formula, which specifies landslide occurrence (`lslpts`) as a
    function of the predictors
-   A family, which specifies the type of model, in this case binomial
    because the response is binary (see `?family`)
-   The data frame which contains the response and the predictors (as
    columns)

The results of this model can be printed as follows (summary(fit)
provides a more detailed account of the results):

```{r}
form <- lslpts ~ slope + cplan + cprof + elev + log10_carea

fit <- glm(form, family = binomial(), data = lsl)
summary(fit)
```

The binomial GLM indicates that landslide probability increases
significantly with **slope**, confirming the dominant role of terrain
steepness. **Plan curvature (cplan)** and **contributing area
(log10_carea)** show strong negative effects, suggesting lower
susceptibility on divergent surfaces and in areas with larger upslope
drainage. **Profile curvature (cprof)** and **elevation** are not
statistically significant once other variables are accounted for.
Overall, the reduction in deviance relative to the null model shows that
topographic predictors explain a substantial part of landslide
occurrence, with slope and curvature being the key controls. The
estimated coefficients are expressed in **log-odds**: a positive
estimate means an increase in landslide probability, while a negative
estimate indicates a decrease, all else equal. Exponentiating a
coefficient gives an **odds ratio**, quantifying how the odds of a
landslide change for a one-unit increase in the predictor. For
**slope**, an estimate of **0.079** means that a one-unit increase in
slope increases the **log-odds** of a landslide by 0.079, all else
equal. In terms of odds, this corresponds to an **odds ratio of
exp(0.079) ≈ 1.08**, i.e. about an **8% increase in the odds** of a
landslide per unit increase in slope. The effect is highly significant,
indicating that steeper terrain is consistently associated with higher
landslide susceptibility. Because predictors are on different scales,
coefficient magnitudes should be interpreted relative to their units or
visualized using effect plots rather than compared directly.

### Interpret effects (optional)

```{r}
plot(allEffects(fit))
```

The effect plots show a strong, monotonic increase in landslide
probability with slope, confirming it as the dominant control. Plan
curvature (cplan) and contributing area (log10_carea) exhibit clear
negative effects, indicating lower susceptibility on divergent terrain
and in areas with larger drainage accumulation. Elevation and profile
curvature (cprof) show weak or uncertain effects, consistent with their
non-significant coefficients in the model.

------------------------------------------------------------------------

## 5. Spatial prediction (susceptibility mapping)

The model object `fit` (of class `glm`) contains the coefficients
describing the fitted relationship between response and the predictors.
It can also be use for prediction. For example, we can predict the
landslide probabilities for each of the observed points in the
`lsl`dataframe using the generic `predict`method. Setting the type to
`response`returns the predicted probabilities (of landslide occurence):

```{r}
pred_glm = predict(object = fit, type = "response")
head(pred_glm)
```

This table shows for each of the observed landslides the predicted
probability of landslide occurence. Spatial distribution maps can then
be made by applying the same approach where we use predictor maps rather
than the point observations using the `terra:predict()`method. We
predict probabilities over the raster covariates and display a 5-class
susceptibility map.

```{r}
pred <- terra::predict(ta, model = fit, type = "response")
pred_masked <- mask(pred, study_mask)

breaks <- seq(0, 1, by = 0.2)
labels <- c("Very Low", "Low", "Moderate", "High", "Very High")

pred_class <- classify(pred_masked,
                       rcl = cbind(breaks[-length(breaks)], breaks[-1], 1:5))
levels(pred_class) <- data.frame(id = 1:5, class = labels)

pred_df <- as.data.frame(pred_class, xy = TRUE)
colnames(pred_df) <- c("x", "y", "class")
#pred_df$class <- factor(pred_df$class, levels = 1:5, labels = labels)

ggplot() +
  geom_raster(data = pred_df, aes(x = x, y = y, fill = class)) +
  scale_fill_manual(
    values = c("Very Low" = "#ffffcc",
               "Low"      = "#a1dab4",
               "Moderate" = "#41b6c4",
               "High"     = "#2c7fb8",
               "Very High"= "#253494"),
    name = "Susceptibility"
  ) +
  geom_sf(data = study_mask, fill = NA, color = "black", linewidth = 0.4) +
  coord_sf() +
  theme_minimal() +
  labs(title = "Landslide susceptibility map (5 classes)", x = "Easting (m)", y = "Northing (m)")
```

Note that here we neglect spatial autocorrelation. It is possible to
include spatial autocorrelation structures into models as well as into
predictions, but this is beyond the scope of this course (explore
regression kriging or mixed-effect modelling approaches).

::: callout-note
Spatial distribution mapping is one very important outcome of a model.
Even more important is how good the underlying model is at making them
since a prediction map is useless if the model’s predictive performance
is bad.
:::

## 6. Model evaluation: apparent AUC (optimistic baseline)

One of the most popular measures to assess the predictive performance of
a binomial model is the *Area Under the Receiver Operator Characteristic
Curve (AUROC)*. This is a value between 0.5 and 1.0, with 0.5 indicating
a model that is no better than random and 1.0 indicating perfect
prediction of the two classes. Thus, the higher the AUROC, the better
the model’s predictive power. The following code chunk computes the
AUROC value of the model with `roc()`, which takes the response and the
predicted values as inputs. `auc()` returns the area under the curve.

This evaluates on the same data used for training (optimistic but a
useful baseline).

```{r}
auc_full <- pROC::auc(pROC::roc(lsl$lslpts, fitted(fit), quiet = TRUE))
auc_full
```

Now, that isnt a bad fit! Note however that this is a very optimistic
estimation of the model predictions since we have computed it on the
complete dataset. In the sections below we address this issue.

## 7. simple Hold-out split

As a first step beyond apparent performance, we evaluate the model using
a single train–test split to mimic prediction on unseen data. Train on
80% and test on 20% is done here below. While easy to implement, a
single hold-out split offers only a coarse estimate of model performance
and serves mainly as a baseline because it depends on the particular
split.

![Figure: Train/Test Split illustration. Concept adapted from: DataCamp
(n.d.), “K-Fold Cross Validation”.
https://www.datacamp.com/tutorial/k-fold-cross-validation](images/Train_test.png){fig-align="center"
width="80%"}

```{r}
set.seed(1)

n <- nrow(lsl)
id_train <- sample(seq_len(n), size = 0.8 * n)

train <- lsl[id_train, ]
test  <- lsl[-id_train, ]

fit_split <- glm(form, family = binomial(), data = train)

p_test  <- predict(fit_split, newdata = test, type = "response")
p_train <- fitted(fit_split)

auc_test  <- pROC::auc(pROC::roc(test$lslpts,  p_test,  quiet = TRUE))
auc_train <- pROC::auc(pROC::roc(train$lslpts, p_train, quiet = TRUE))

cat("AUC (training dataset):", round(auc_train, 3), "\n")
cat("AUC (testing dataset): ", round(auc_test,  3), "\n")
```

The higher AUC on the training set reflects optimistic in-sample
performance, while the test AUC provides a more realistic estimate of
predictive skill on unseen data.

------------------------------------------------------------------------

## 8. k-fold cross-validation (random CV)

Building on the simple hold-out split, k-fold cross-validation provides
a more robust estimate of predictive performance by repeating the
train–test procedure multiple times. The data are partitioned into *k*
subsets, each of which is used once for testing while the remaining
folds are used for training. By averaging performance across folds,
k-fold cross-validation reduces the dependence on a single random split
and yields a more stable assessment of model skill.

::: callout-note
In standard k-fold CV:

1.  The dataset is randomly partitioned into k equally sized subsets.
2.  Each subset is used once as a test set.
3.  Performance is averaged across folds.

This estimates the expected prediction error under the assumption that
future data come from the *same joint distribution* as the observed
data.
:::

![Figure: Principle of cross-validation. The average performance for the
5 models than represents the overall performance after cross-validation
(source:
https://www.datacamp.com/tutorial/k-fold-cross-validation).](https://media.datacamp.com/legacy/v1718738336/image_0bb32b40f1.jpg)

In R, we can implement this functionality as follows:

```{r}
auc01 <- function(y, p) as.numeric(pROC::auc(pROC::roc(y, p, quiet = TRUE)))

set.seed(1)
k <- 5
fold_id <- sample(rep(1:k, length.out = nrow(lsl)))

auc_kfoldcv <- numeric(k)

for (i in 1:k) {
  train_i <- lsl[fold_id != i, ]
  test_i  <- lsl[fold_id == i, ]

  fit_i <- glm(form, family = binomial(), data = train_i)
  p_i   <- predict(fit_i, newdata = test_i, type = "response")

  auc_kfoldcv[i] <- auc01(test_i$lslpts, p_i)
}

data.frame(
  Fold = seq_along(auc_kfoldcv),
  AUC  = round(auc_kfoldcv, 3)
)
cat(round(mean(auc_kfoldcv), 3), "\n")
```

------------------------------------------------------------------------

## 9. Forms of cross-validation

There are many forms of cross-validation. **Stratified
cross-validation** extends standard k-fold CV by ensuring that each fold
preserves the class proportions of the full dataset. This is
particularly important in classification problems with imbalanced
classes, as it prevents folds that contain too few (or no) positive
cases. Stratification improves the stability and interpretability of
performance metrics such as AUC.

**Group cross-validation** is used when observations are not independent
but belong to meaningful groups (e.g. sites, subjects, time series, or
spatial units). Entire groups are assigned to either the training or
testing set, preventing information leakage across folds. This approach
provides a more realistic assessment of predictive performance when the
goal is to generalize to new groups rather than new observations within
the same group.

| Type of Cross-Validation       | Typical Usage                                 | Description                                                                                        | When to Use                                                                          |
|--------------------------------|-----------------------------------------------|----------------------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------|
| **Standard k-Fold CV**         | Regression & Classification                   | Splits the dataset into *k* equal-sized folds; each fold is used once as the test set.             | Best for balanced datasets to obtain a stable overall performance estimate.          |
| **Stratified k-Fold CV**       | Classification                                | Preserves the class proportions of the original dataset within each fold.                          | Recommended for imbalanced classification problems.                                  |
| **Leave-One-Out CV (LOOCV)**   | Regression & Classification                   | Each observation is left out once as the test set, with all remaining data used for training.      | Suitable for very small datasets; computationally expensive and often high-variance. |
| **Leave-P-Out CV**             | Regression & Classification                   | Similar to LOOCV, but leaves out *p* observations per iteration.                                   | Useful for small datasets to explore sensitivity to data removal.                    |
| **Group k-Fold CV**            | Regression & Classification with grouped data | Ensures that all observations from the same group are kept together in either training or testing. | Appropriate when observations are not independent (e.g. repeated measures, sites).   |
| **Stratified Group k-Fold CV** | Classification with grouped data              | Combines stratification and group integrity, preserving class balance while respecting groups.     | Ideal for grouped and imbalanced classification problems.                            |

Leave-One-Out Cross-Validation (LOOCV) is an extreme form of k-fold
cross-validation in which each observation is used once as the test set
while all remaining observations are used for training. It maximizes
training data per fit but is computationally expensive and often yields
high-variance performance estimates, especially for spatial data. AUC
cannot be computed per fold (one test observation), so we compute one
pooled AUC from all out-of-fold predictions.

```{r}
p_loocv <- numeric(nrow(lsl))

for (i in 1:nrow(lsl)) {
  fit_i <- glm(form, family = binomial(), data = lsl[-i, ])
  p_loocv[i] <- predict(fit_i, newdata = lsl[i, , drop = FALSE], type = "response")
}

auc_loocv <- pROC::auc(pROC::roc(lsl$lslpts, p_loocv, quiet = TRUE))
auc_loocv
```

------------------------------------------------------------------------

## 10. Spatial CV

Random CV assumes observations are (approximately) independent. Spatial
data often violate this due to spatial autocorrelation, leading to
over-optimistic performance estimates for spatial prediction. An
excellent source that introduces this topic is @Sun2023SpatialCV. The
key is to split the data spatially rather than randomly, thereby
increasing the independence between training and validation data. There
are several different types of spatial CV and here we introduce the main
methods.

::: callout
Spatial CV changes the validation question from: How well do I predict
new *observations*? to: How well do I predict at *new locations*? This
yields a more realistic estimate for spatial generalization.
:::

------------------------------------------------------------------------

### Spatial CV: cluster-based folds

Cluster-based spatial CV separates data into k spatial clusters and uses
them as folds. One method to split data spatially is to perform spatial
clustering. Based on the geospatial coordinates of the data and
leveraging a clustering algorithm (e.g., K-means clustering), we can
split data into several subsets with data instances within the same
subset being spatially continuous and those in different subsets being
relatively separated from each other. We can do this using the `blockCV`
package.

```{r}
library(blockCV)

lsl_sf_utm <- st_as_sf(lsl, coords = c("x", "y"), crs = 32717)

set.seed(1)
cv_clust <- blockCV::cv_cluster(
  x      = lsl_sf_utm,
  column = "lslpts",
  k      = 5
)

fold_id_sp <- cv_clust$folds_ids
table(fold_id_sp)
```

```{r}
ksp <- length(unique(fold_id_sp))
cols <- grDevices::hcl.colors(ksp, palette = "Dark 3")
pt_cols <- cols[as.integer(factor(fold_id_sp, levels = sort(unique(fold_id_sp))))]

plot(st_geometry(lsl_sf_utm), pch = 16, cex = 0.9, col = pt_cols,
     main = "Cluster-based spatial CV folds (cv_cluster)")
legend("topright",
       legend = paste("Fold", sort(unique(fold_id_sp))),
       col = cols, pch = 16, pt.cex = 0.9, bty = "n")
```

### Evaluate GLM with cluster-fold CV

```{r}
auc_cluster <- numeric(length(unique(fold_id_sp)))

for (i in sort(unique(fold_id_sp))) {
  train_sf <- lsl_sf_utm[fold_id_sp != i, ]
  test_sf  <- lsl_sf_utm[fold_id_sp == i, ]

  train <- st_drop_geometry(train_sf)
  test  <- st_drop_geometry(test_sf)

  fit_i <- glm(form, family = binomial(), data = train)
  p_i   <- predict(fit_i, newdata = test, type = "response")

  auc_cluster[i] <- auc01(test$lslpts, p_i)
}

auc_cluster
mean(auc_cluster)
```

Now that is all good, but how do we know how many blocks we should
choose?

### Choosing a block size from spatial autocorrelation (principle)

Block-based spatial cross-validation should be informed by the **spatial
range of autocorrelation** present in the data. By analysing spatial
autocorrelation in model residuals (e.g. using a correlogram or Moran’s
I over distance classes), one can identify the distance beyond which
observations become approximately independent. Choosing a block size
equal to or slightly larger than this range reduces information leakage
between training and test sets and yields a more realistic estimate of
predictive performance at new locations. We estimate the spatial
dependence range in **model residuals** (dependence left after
accounting for predictors). Below: Moran’s I correlogram of Pearson
residuals over distance bands.

::: callout-note
A key insight is that **block size should reflect spatial
autocorrelation**. The workflow used here follows a principled logic: 1.
Fit a model. 2. Analyze spatial autocorrelation in the residuals. 3.
Identify the distance at which residual autocorrelation vanishes. 4. Use
that distance (or slightly larger) as the block size. Residual-based
autocorrelation matters because: it represents structure not explained
by the model, and therefore the information that may leak between nearby
training and test points.
:::

```{r}
library(spdep)

fit_all <- glm(form, family = binomial(), data = st_drop_geometry(lsl_sf_utm))
lsl_sf_utm$pearson_resid <- residuals(fit_all, type = "pearson")

coords <- st_coordinates(lsl_sf_utm)

D <- as.matrix(dist(coords))
dmax <- as.numeric(quantile(D[upper.tri(D)], 0.95))

nbins <- 15
breaks <- seq(0, dmax, length.out = nbins + 1)

moran_I <- numeric(nbins)
moran_p <- numeric(nbins)
mid_d   <- numeric(nbins)

for (b in 1:nbins) {
  d1 <- breaks[b]
  d2 <- breaks[b + 1]
  mid_d[b] <- (d1 + d2) / 2

  nb <- dnearneigh(coords, d1 = d1, d2 = d2, longlat = FALSE)
  lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

  mt <- moran.test(lsl_sf_utm$pearson_resid, lw, zero.policy = TRUE)
  moran_I[b] <- mt$estimate[["Moran I statistic"]]
  moran_p[b] <- mt$p.value
}

corr_tbl <- data.frame(distance_m = mid_d, moran_I = moran_I, p_value = moran_p)
corr_tbl
```

```{r}
plot(mid_d, moran_I, type = "b", pch = 16,
     xlab = "Distance (m)", ylab = "Moran's I (Pearson residuals)",
     main = "Residual spatial autocorrelation (Moran correlogram)")
abline(h = 0, lty = 2)
sig <- moran_p < 0.05
points(mid_d[sig], moran_I[sig], pch = 16, cex = 1.3)
```

------------------------------------------------------------------------

### Spatial CV: Grid-based

Apply block CV with a chosen `size` (meters, EPSG:32717). If the study
area is small, a large block size may yield too few blocks for k folds.

```{r}
block_size_m <- 1500  # adjust based on correlogram range and feasibility

set.seed(1)
cv_block <- blockCV::cv_spatial(
  x = lsl_sf_utm,
  column = "lslpts",
  k = 8,
  size = block_size_m,
  selection = "random",
  iteration = 500,
  progress = FALSE
)

fold_id_block <- cv_block$folds_ids
table(fold_id_block)
```

### Evaluate GLM with block-fold CV

```{r}
kblk <- length(unique(fold_id_block))
auc_spatial <- numeric(kblk)

for (i in sort(unique(fold_id_block))) {
  train_sf <- lsl_sf_utm[fold_id_block != i, ]
  test_sf  <- lsl_sf_utm[fold_id_block == i, ]

  train <- st_drop_geometry(train_sf)
  test  <- st_drop_geometry(test_sf)

  fit_i <- glm(form, family = binomial(), data = train)
  p_i   <- predict(fit_i, newdata = test, type = "response")

  auc_spatial[i] <- auc01(test$lslpts, p_i)
}

auc_spatial
mean(auc_spatial)
```

### Visualize blocks + fold membership

```{r}
cols_blk <- grDevices::hcl.colors(kblk, palette = "Dark 3")

plot(st_geometry(cv_block$blocks), border = "grey70",
     main = paste0("Block CV folds (cv_spatial), size = ", block_size_m, " m"))
plot(st_geometry(lsl_sf_utm), add = TRUE, pch = 16, cex = 0.8, col = cols_blk[fold_id_block])

legend("topright", legend = paste("Fold", sort(unique(fold_id_block))),
       col = cols_blk, pch = 16, bty = "n")
```

------------------------------------------------------------------------

### Summary: compare evaluation strategies

```{r}
summary_tbl <- data.frame(
  method = c("Apparent (train=test)", "Hold-out (test AUC)", "Random k-fold CV (mean AUC)",
             "LOOCV (pooled AUC)", "Spatial CV (cluster mean AUC)", "Spatial CV (block mean AUC)"),
  AUC = c(as.numeric(auc_full),
          as.numeric(auc_test),
          mean(auc_kfoldcv),
          as.numeric(auc_loocv),
          mean(auc_cluster),
          mean(auc_spatial))
)

summary_tbl
```

::: callout-note
## 11. Take-home messages & Questions

-   Cross-validation estimates performance under specific assumptions.
-   Random CV assumes independence and exchangeability.
-   Spatial CV changes the estimand to spatial generalization error.
-   Block size should be informed by residual autocorrelation, not
    convenience.
:::

------------------------------------------------------------------------

### 💡Questions

> **Question:** What assumption does random k-fold CV make about the
> data?

::: {.callout-tip collapse="true"}
Random CV assumes observations are exchangeable. In spatial data, this
is often violated because nearby observations are more similar than
distant ones.
:::

> **Question:** Why is AUC computed on the training data misleading?

::: {.callout-tip collapse="true"}
Because the model is evaluated on the same data used to fit it, leading
to overly optimistic performance that does not reflect prediction at new
locations.
:::

> **Question:** What question does spatial CV answer that random CV does
> not?

::: {.callout-tip collapse="true"}
Spatial CV estimates performance at *new locations*, not just new
observations close to the training data.
:::

> **Question:** Why is spatial autocorrelation assessed on model
> residuals?

::: {.callout-tip collapse="true"}
Residuals represent unexplained structure. Autocorrelation in residuals
indicates spatial dependence that can leak information between training
and test data.
:::

## 12. Tidymodels: formalising the workflow.

The above examples work well but require alot of coding in basic R and
are not always transposable to other type of models. To address this, we
can work with modelling pipeline that creates a uniform interface for
the massive variety of statistical and machine learning models that
exist. Here we are going to work with `tidymodels`. Tidymodels is a
**suite of R packages** designed to make modeling workflows
**consistent, modular, and reproducible**.\
Instead of mixing preprocessing, model fitting, and validation in ad-hoc
code, tidymodels encourages you to define:

-   a **recipe** (preprocessing steps),
-   a **model specification** (algorithm + engine),
-   a **workflow** (recipe + model combined),
-   and a **resampling plan** (e.g., k-fold CV) for evaluation and
    (later) tuning.

This section follows the logic described by Rebecca Barter’s
introduction to tidymodels.\
(Reference:
https://rebeccabarter.com/blog/2020-03-25_machine_learning#what-is-tidymodels)

### 12.1 Install/load

```{r, ECHO = FALSE}
library(tidymodels)
library(tidyverse)
library(workflows)
```

### 12.2 Train/Test split

Let's first split the dataset we were working with `lsl`into training
and testdata. This split can be done automatically using the
`initial_split`function from `rsample`. Note that we use
`strata`argument to sample randomly within the stratification variable
(`lslpts`).

```{r}
set.seed(234589)

# 75% train, 25% test

lsl_split <- initial_split(lsl, prop = 3/4, strata = lslpts)

lsl_train <- training(lsl_split)
lsl_test  <- testing(lsl_split)

lsl_split

# 1) Create k-fold resamples (usually from the training set)
set.seed(1)
lsl_folds <- vfold_cv(lsl_train, v = 5, strata = lslpts)
lsl_folds

```

The output of `lsl_split` is an object that contains the split
information as well as the original dataframe.

### 12.3 Recipe

Recipes allow you to specify the role of each variable as an outcome or
predictor variable (using a “formula”), and any pre-processing steps you
want to conduct (such as normalization, imputation, PCA, etc).

Creating a recipe has two parts (layered on top of one another using
pipes `%>%`):

1.  **Specify the formula** (`recipe()`): specify the outcome variable
    and predictor variables

2.  **Specify pre-processing steps** (`step_zzz()`): define the
    pre-processing steps, such as imputation, creating dummy variables,
    scaling, and more.

```{r}
lsl_recipe <- recipe(lslpts ~ slope + cplan + cprof + elev + log10_carea,
data = lsl_train) %>%
step_normalize(all_numeric_predictors()) %>%     # scale/center predictors
step_nzv(all_predictors())                       # remove near-zero variance predictors

lsl_recipe

```

Notice how we define the model
`lslpts ~ slope + cplan + cprof + elev + log10_carea`in the recipy,
define the data, and the pre-processing steps that we want. You can find
a full description of potential pre-processing steps here
<https://recipes.tidymodels.org/index.html>. Note that we use the
functions `` all_numeric_predictors` and all_predictors `` to specify
and what variables you want to apply the pre-processing. You can explore
this further by using `?selections`.

::: callout-note
It is important to realize that nothing has happened so far, we only
defined the names and roles of the outcome and predictor variables in
this recipe. We used `lsl_train` as dataset, we we could have also used
`lsl`. We will apply this recipe later to datasets.
:::

On to the next step, the model

## 12.4 Specify the model

At this stage, we explicitly define the statistical model and connect it
to the data preprocessing steps in a single, reproducible workflow. We
specify a logistic regression model (see above, we will work on the same
landslide example), which is appropriate because the response variable
is binary and the objective is to estimate the probability of occurrence
of the target process.

Using a workflow ensures that the model and the preprocessing recipe are
treated as a single unit. This prevents data leakage, guarantees that
the same transformations are applied consistently during resampling and
prediction, and improves reproducibility. By separating what the model
is (logistic regression) from how the data are prepared (the recipe),
the workflow provides a transparent and extensible structure for
evaluation, cross-validation, and spatial prediction.

There are a few primary components that you need to provide for the
model specification

-   The model type: what kind of model you want to fit, set using a
    different function depending on the model, such as rand_forest() for
    random forest, logistic_reg() for logistic regression, svm_poly()
    for a polynomial SVM model etc. The full list of models available
    via parsnip can be found
    [here](https://www.tidymodels.org/find/parsnip/).

-   The arguments: the model parameter values (now consistently named
    across different models), set using set_args(). We will deal with
    this later.

-   The engine: the underlying package the model should come from (e.g.
    “glm” for the logistic_reg), set using set_engine().

-   The mode: the type of prediction - since several packages can do
    both classification (binary/categorical prediction) and regression
    (continuous prediction), set using set_mode()

In our test case we need:

```{r}
lr_model <- logistic_reg() %>% set_engine("glm") %>% set_mode("classification")


```

::: callout-note
Note that this code doesn’t actually fit the model. Like the recipe, it
just outlines a description of the model. Also note that we do not deal
with tuning the model at this stage (see later).
:::

## 12.5 Putting it all togetheter in a workflow

Now we can put the model and recipe together into a worklfow. This can
be done by using `workflow()`(from the `workflows`package) and by adding
the recipe and model to it. Nice!

```{r}
lr_workflow <- workflow() %>% add_recipe(lsl_recipe) %>% add_model(lr_model)

lr_workflow


```

## 12.6 Evaluate the model

We can now evaluate the workflow using two options; - evaluate the
workflow using a held-out set (as we defined by
`lsl_split') -  or evaluate using a k-fold cross validation (as we defined by 'lsl_folds`)

For the first, use `last_fit`, for the cross-validation, use
`fit_resamples`

```{r}
lr_fit <- lr_workflow %>%
  last_fit(
    split   = lsl_split,
    metrics = metric_set(roc_auc)
  )
lr_fit %>%
  collect_metrics()
```

```{r}
lr_cv_fit <- lr_workflow %>%
  fit_resamples(
    resamples = lsl_folds,
    metrics   = metric_set(roc_auc)
  )
lr_cv_fit %>%
  collect_metrics()
```
