--- 
title: "Solution Challenge Session 4 2024"
author: Kristof Van Oost, Valentin 
date: March 2024
output: 
  html_document: 
    # toc: yes
    toc: true
    # theme: united
    toc_depth: 3
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

# 1. Objective
The objective of our geoproject was to map the Habitat distribution of a given species, in a given area, In a given temporal moment: current and future climatic conditions. The output should be a change analysis current -> to future (2050 -> 2070 etc)... Something like this
```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(geodata)
library(terra) 
library(raster)
library(dplyr)
#setwd("/Users/kvanoost/Downloads/Session 5")
options(geodata_default_path = "/Users/kvanoost/Documents/UCL/Cours/GIS/geodata")
suitability_map_fct<-function(region,genus,species,res,ssp,model){
  ###Description: Function to model the suitability (present and future) of an area for a species.
  ###Inputs:
  ##region of interest (spatVector) should be a country with ISO code , 
  ##genus and species of interset (see https://www.gbif.org/)
  ##res (numeric): resolution 2.5 5 or 10, 
  ##ssp : (character). A valid Shared Socio-economic Pathway code: "126", "245", "370" or "585"., 
  ##model: (character) climatic model for future prediction (see http://worldclim.com/CMIP5v1)
  #The possible CMIP6 models are Climate model abbrevation. One of "ACCESS-CM2", "ACCESS-ESM1-5", "AWI-CM-1-1-MR", "BCC-CSM2-MR", "CanESM5", "CanESM5-CanOE", "CMCC-ESM2", "CNRM-CM6-1", "CNRM-CM6-1-HR", "CNRM-ESM2-1", "EC-Earth3-Veg", "EC-Earth3-Veg-LR", "FIO-ESM-2-0", "GFDL-ESM4", "GISS-E2-1-G", "GISS-E2-1-H", "HadGEM3-GC31-LL", "INM-CM4-8", "INM-CM5-0", "IPSL-CM6A-LR", "MIROC-ES2L", "MIROC6", "MPI-ESM1-2-HR", "MPI-ESM1-2-LR", "MRI-ESM2-0", "UKESM1-0-LL"
  ###Output: 3 suitability maps (or corresponding RasterLayers)
  
  
  ##1. Load sightings data and verification
  
  #1.a Download observations
  sightings <- sp_occurrence(genus=genus,species = species,ext=(region), sp=TRUE, download = TRUE, geo=TRUE)
  sightings <- vect(sightings)
  crs(sightings) <- crs(region)
  
  #1.b Check if there're enough observations of the species in ROI
  if (length(sightings) < 40) stop("Not enough sightings (ie < 40) of this species in the selected region of interest to model the suitability map. Try another species and/or area of study.")
  
  
  ##2. create RasterStack with predictors
  mybio <- worldclim_global(var='bio', res=res,ssp=ssp) #https://worldclim.org/data/bioclim.html
  mybio_aoi <- mybio %>% crop(region) %>% mask(region)
  
  ##3. Create presence/abscence database
  
  sightings$pres<-1
  
  #3.a Generating random absence points
  set.seed(1975) 
  abs <-  spatSample(region,size=length(sightings),method="random")
  abs$pres=0
  
  #3.b merge the sightings and absence points
  my_occ_df=sightings+abs
  
  #3.c Extract bio variables (from mybio_aio) values for each observation
  my_train_df<-terra::extract(mybio_aoi,my_occ_df,df=TRUE,bind=TRUE)
  my_train_df <- as.data.frame(my_train_df)
  my_train_df <- my_train_df %>% dplyr::select(pres, starts_with("wc")) 
  
  
  ##4. Modelling
  #4.a Exploratory analysis 
  # summary(my_train_df)
  # correlation<-cor(my_train_df, use="complete.obs") #there are some NA
  # library(corrplot)
  # corrplot(correlation)
  
  #4.b model
  myMod<-glm(pres~ . , data=my_train_df, family = "binomial")
  # summary(myMod)
  # #check the imporance of the variables: dropterm function removes one-by-one one of the variable and test its imporance in the model selection procedure. 
  # # library(MASS)
  # # dropterm(myMod, test = "Chisq")
  # library(modEvA)
  # Dsquared(myMod, adjust = TRUE)
  
  
  ##5. Predict habitat occurence

  #5.b Creating suitability layers
  library(leaflet)
  library(dplyr)
  a<-c("1970-2000","2021-2040", "2041-2060",  "2061-2080")
  pal <- colorNumeric(c("#FFFFCC", "#41B6C4","#0C2C84" ), domain = c(0,1),
                      na.color = "transparent")
  ####### Setup loop
  
  years<-c("1970-2000","2021-2040", "2041-2060",  "2061-2080")
  
  my_list <- list() 
  
  for(j in 1:length(years)) {                 # Head of for-loop
    if (years[j]=="1970-2000"){
      raster<-mybio_aoi
    } else {
      #raster<-raster::getData('CMIP5', var='bio', res=res, rcp=rcp, model=model, year= as.numeric(a[j]))
      raster <- cmip6_world(model=model, ssp=ssp, time=years[j], var="bioc", res=res)
      raster<-crop(mask(raster, region),region)
      names(raster)<-names(mybio_aoi)
    }       
    
  
    Pred<-predict(raster,myMod,type='response')   ### Rastermap
    
    map<-leaflet::leaflet() %>% leaflet::addTiles() %>%
      leaflet::addRasterImage(raster(Pred), colors = pal, opacity = 0.8) %>%
      leaflet::addLegend(pal = pal, values = raster::values(Pred),
                         title = paste("Suitability",a[j]))
    output<-list(Pred,map)
    my_list[[j]] <- output        # Store output, which is a list, in another list
  }
  ######
  
  #############
  
  #6. Creation outputs
  result <- my_list
  maps<-list(result[[1]][[2]],result[[2]][[2]],result[[3]][[2]],result[[4]][[2]])
  names(maps) <- c('suitabilitymap_2000','suitabilitymap_2030','suitabilitymap_2050','suitabilitymap_2070')
  layers<-list(result[[1]][[1]],result[[2]][[1]],result[[3]][[1]],result[[4]][[1]])
  names(layers) <- c('suitabilityraster_2000','suitabilityraster_2030','suitabilityraster_2050','suitabilityraster_2070')
  
  #6.a Change analysis
  change30<-layers[[2]]-layers[[1]]
  change50<-layers[[3]]-layers[[1]]
  change70<-layers[[4]]-layers[[1]]
  
  pal <- colorNumeric(c("#c45741","#FFFFCC", "#0C2C84" ), domain = c(-1,1),
                      na.color = "transparent")
  map_change30<-leaflet::leaflet() %>% leaflet::addTiles() %>%
    leaflet::addRasterImage(raster(change30), colors = pal, opacity = 0.8) %>%
    leaflet::addLegend(pal = pal, values = values(change30),
                       title = paste("Suitability change for 2030"))
  map_change50<-leaflet::leaflet() %>% leaflet::addTiles() %>%
    leaflet::addRasterImage(raster(change50), colors = pal, opacity = 0.8) %>%
    leaflet::addLegend(pal = pal, values = values(change50),
                       title = paste("Suitability change for 2050"))
  map_change70<-leaflet::leaflet() %>% leaflet::addTiles() %>%
    leaflet::addRasterImage(raster(change70), colors = pal, opacity = 0.8) %>%
    leaflet::addLegend(pal = pal, values = values(change70),
                       title = paste("Suitability change for 2070"))
  maps_change <- list(map_change30,map_change50,map_change70)
  names(maps_change) <- c('changemap_2030','changemap_2050','changemap_2070')
  #7. Return
  #return(list(layers,list(change50,change70,change50_70)))
  results<-(list(maps,maps_change,layers))
  names(results) <- c('suitability_map','change_map','raster_layers')
  return(results)
  #output : a list the 3 predictions and the 3 change rasterlayers or the same as interactive maps
}

region <- gadm(country="MEX", level=0, path='GADM')

output<-suitability_map_fct(region=region,genus = 'ambystoma', species = 'mexicanum',res=10,model="ACCESS-CM2",ssp="126")

library(leafsync)
sync(output$suitability_map)

```

# 2. Functionality

> In a first step, you can define the required functionality by setting up your function and by defining the variables and required output. Not

```{r, eval=FALSE}
suitability_map_fct<-function(region,genus,species,res,ssp,model){
  ###Description: Function to model the suitability (present and future) of an area for a species.
  ###Inputs:
  ##region of interest (spatVector) should be a country with ISO code , 
  ##genus and species of interest (see https://www.gbif.org/)
  ##res (numeric): resolution 2.5 5 or 10, 
  ##ssp : (character). A valid Shared Socio-economic Pathway code: "126", "245", "370" or "585"., 
  ##model: (character) climatic model for future prediction (see http://worldclim.com/CMIP5v1)
  #The possible CMIP6 models are Climate model abbrevation. One of "ACCESS-CM2", "ACCESS-ESM1-5", "AWI-CM-1-1-MR", "BCC-CSM2-MR", "CanESM5", "CanESM5-CanOE", "CMCC-ESM2", "CNRM-CM6-1", "CNRM-CM6-1-HR", "CNRM-ESM2-1", "EC-Earth3-Veg", "EC-Earth3-Veg-LR", "FIO-ESM-2-0", "GFDL-ESM4", "GISS-E2-1-G", "GISS-E2-1-H", "HadGEM3-GC31-LL", "INM-CM4-8", "INM-CM5-0", "IPSL-CM6A-LR", "MIROC-ES2L", "MIROC6", "MPI-ESM1-2-HR", "MPI-ESM1-2-LR", "MRI-ESM2-0", "UKESM1-0-LL"
  ###Output: 3 suitability maps (or corresponding RasterLayers)
  }
```


# 3. Create your predictive model

This part is exactly the same as the first part of the project from Session 4. I repeat the steps we took last week:

    1. Load sightings data and verification
       
       -1.a Download observations
       -1.b Check if there're enough observations of the species in ROI
       
    2. create RasterStack with predictors
    
    3. Create presence/abscence database
    
       -3.a Generating random absence points
       -3.b merge the sightings and absence points
       -3.c Extract bio variables (from mybio_aio) values for each observation
       
    4. Modelling
    
       -4.a Exploratory analysis 
       -4.b model

So you end with this model, which is built on the present-day observations and climatic conditions

>  `myMod<-glm(pres~ . , data=my_train_df, family = "binomial")`

# 4. Create maps for future climate  

Start with creating a vector with the years (compatible with the geodata function)
  
  > `years<-c("1970-2000","2021-2040", "2041-2060",  "2061-2080")`

Create an empty list to store the results

>  `my_list <- list()` 

The main loop
```{r, eval=FALSE, warning=FALSE, message=FALSE}
for (j in 1:length(years)) {
 if (years[j]=="1970-2000") { mybio <- raster_you_defined_above for current conditions }
      else 
     {
      raster <- cmip6_world(model=model, ssp=ssp, time=years[j], var="bioc", res=res)
      raster<-crop(mask(raster, region),region)
      names(raster)<-names(mybio_aoi)
     }      
  Pred<-predict(raster,myMod,type='response')   ### Rastermap
  map<-leaflet. create_your_map
  output<-list(Pred,map). #Store the map and the rasters in a list
  my_list[[j]] <- output        # Store output, which is a list, in another list
  }
     
```

Finish with creating some easy-to-use names in your list

```{r, eval=FALSE}
  result <- my_list
  maps<-list(result[[1]][[2]],result[[2]][[2]],result[[3]][[2]],result[[4]][[2]])
  names(maps) <- c('suitabilitymap_2000','suitabilitymap_2030','suitabilitymap_2050','suitabilitymap_2070')
  layers<-list(result[[1]][[1]],result[[2]][[1]],result[[3]][[1]],result[[4]][[1]])
  names(layers) <- c('suitabilityraster_2000','suitabilityraster_2030','suitabilityraster_2050','suitabilityraster_2070')
  
  result <- list(map,layers)
```
  
# Example

```{r, warning=FALSE, message=FALSE}
#Example using the fct

## Define your variables 
region <- gadm(country="MEX", level=0, path='GADM')

output<-suitability_map_fct(region=region,genus = 'ambystoma', species = 'mexicanum',res=10,model="ACCESS-CM2",ssp="126")

```

Note the easy use of output:
```{r, warning=FALSE, message=FALSE}
output$suitability_map$suitabilitymap_2030
output$raster_layers$suitabilityraster_2050
```

  
