---
title: "Session 3 : Geoprocessing with vectors"
author: "Kristof Van Oost & Valentin Charlier"
Date: 21/02/2024
theme: zephyr
toc-title: "Geoprocessing with vectors : Toc"
toc: true
toc-depth: 5
standalone: true
embed-resources: true
format: html
editor: visual
execute:
  echo: True
  warning: false
  error: false
---

```{=html}
<style>
body {
text-align: justify}
</style>
```
## Today's program

-   Last week's assignment: discussion
-   Geoprocessing
-   Exercise all together

## Today's Learning Objectives

-   Knowing how to manipulate spatial object in R

## Additional resources

-   The book "Geocomputation in R" (<https://geocompr.robinlovelace.net/>)
-   sf pkg \[vignette\] (<https://cran.r-project.org/web/packages/sf/vignettes/sf1.html>)
-   Spatial data manipulation in [R](https://www.rspatial.org/)
-   Example on spatial data manipulation \[here\] (<https://fukamilab.github.io/BIO202/08-A-mapping.html>) and \[here\] (<https://nceas.github.io/oss-lessons/spatial-data-gis-law/3-mon-intro-gis-in-r.html>)
-   Plots your data using [ggplot2](http://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html)
-   dataframes manipulation using [reshape2](https://seananderson.ca/2013/10/19/reshape/)

## Geoprocessing

```{r, eval=FALSE}
install.packages("sp")
install.packages("raster")
install.packages("terra")
install.packages("geodata")
install.packages("sf")
```

Let's set our working directory and load the packages into the global environment.

```{r, message=FALSE}
setwd("C:/Users/valcharlier/Desktop/Assistanat Géographie/Q2/LGEO2185 Géoprocessing/TP3")
library(sp) #vector
library(terra) # Update of raster
library(geodata)# Update of getData
library(sf)#simple features
library(raster)
```

We will use the geodata::gadm function from the geodata package. Type ?geodata to see the help for this function of the geodata package. We can now easily get data from different online sources such as the Global Administrative Area Data base.

We can get data using the geodata function from the geodata package. Type help(geodata) to see what is possible. Let's download a Spatial Vector DataFrame:

```{r, eval=TRUE, message=FALSE}
France_SVDF <- geodata::gadm("GADM", country = "FRA", level = 2, download = TRUE)
```

In R we can do all the geoprocessing operations that we can normally do in a GIS standalone software.

**Dissolve** First of all we can aggregate, or "dissolve" in GIS jargon. Using the function aggregate() we dissolve all the polygons of the France_SVDF object accordingly the common field 'NAME_1'.

```{r, echo=FALSE, message=FALSE}
pa <- aggregate(France_SVDF, by='NAME_1')
```

```{r}
plot(pa)
```

**Difference** Let's now load a general polygon (available in the "Session 3" Materials folder )

```{r}
pol<-vect("data/fr_poly.shp")
pol
plot(France_SVDF)
plot(pol, add=T, col="blue")
```

We can then compute the difference between two polygons

```{r}
#erase or difference
e <- erase(France_SVDF, pol)
# equivalent to
e <- France_SVDF -pol  
plot(e)
```

**Intersect** The intersection

```{r}
i <- intersect(France_SVDF, pol)
plot(i)

```

**Union** The union the function only works with a few tricks! First, convert our *spatial vector* objects to simple features *sf* and use the function *st_as_s2* if FALSE, polygons that cover more than half the globe are inverted.

```{r}
pol_sf<-st_as_sf(pol)
Fra_sf<-st_as_sf(France_SVDF)

sf_use_s2(FALSE)
U<-st_union(Fra_sf,pol_sf)
plot(U)
plot(U["NAME_1"])
```

**Join** We can non have a look to the join. We first create a fake temperature variable, with one observation for each polygons of the France_SVDF object and we store it in a data frame.

```{r}
nrow(France_SVDF)
temp<-rnorm(n = 96, mean = 25, sd = 4)
temp_df<-data.frame("HASC_2"=France_SVDF$HASC_2, "temp"=temp)
```

We do the join using the merge() function and specifying the common field between the two object.

```{r}
fraL2_temp<-merge(France_SVDF, temp_df, by="HASC_2")
names(France_SVDF)
names(fraL2_temp)
```

**Plot** We are ready now to plot it

```{r}
plot(fraL2_temp, col=fraL2_temp$temp)
```

But we can do something better

```{r}
class(fraL2_temp)
spplot(fraL2_temp, "temp", main = "FRANCE", sub = "mean Temperatures", col = "transparent")
```

But we can improve the color palette and use *spplot* function (Be carefull it does not work all the time with *SpatVector* objects)

```{r}
library(RColorBrewer)
display.brewer.all()
my.palette <- brewer.pal(n = 7, name = "OrRd")

library(classInt)
quantile(fraL2_temp$temp)
breaks.qt <- classIntervals(fraL2_temp$temp, n = 6, style = "quantile", intervalClosure = "right")
spplot(fraL2_temp, "temp", main = "FRANCE", sub = "mean Temperatures", col = "transparent",  col.regions = my.palette, 
       at = breaks.qt$brks)
```

With **terra** you can use the *plot* function to make maps. See ?plot and choose *terra* to get more informations. You can plot a quantitative map using *continuous*,*classes* or *interval* type.

```{r}
my.palette <- brewer.pal(n = 9, name = "YlOrRd")

#Using terra function plot with interval  
plot(fraL2_temp,"temp",type="interval",breaks=6,breakby="cases", col=my.palette,main="Mean temperatures (°) in France")


#Using terra function plot with continuous values 
plot(fraL2_temp,"temp",type="continuous", col=my.palette,main="Mean temperatures (°) in France")

```

### Geoprocessing with "sf"

Have a look to Chapter 4-5 of the book [Geocomputation in R](https://geocompr.robinlovelace.net/) and this [vignette](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html). You can do similar geoprocessing operations using the functions "st_union", "st_join", "st_intersection". Try to understand how to drop the geometrical aspects and how to extract the centroids of a polygon, etc. Usually, making geoprocessing operations with "sf" is faster than using "sp" pkg. Moreover, "sf" pkg as the advantage to be interoperational with the tidyverse packages such as "dplyr" (manipulate data) and "ggplot2" (nice plots!), and you can connect different functions using the "pipe" operator. See a quick example below.

```{r, message=FALSE}
library(sf)
library(ggplot2)
sf::sf_use_s2(TRUE)
France_sf=st_as_sf(France_SVDF)
pol_sf= st_make_valid(st_as_sf(pol, crs=4326))
class(France_sf)

France_sf %>% 
  st_intersection(pol_sf) %>% 
  ggplot()+
  geom_sf(data = France_sf, fill=NA)+
  geom_sf()+
  theme_light()

```

## Exercise 1

1)  Download the data of BELGIUM Level 2
2)  Load landuse file "be_osm_lu_red_v3.shp". The file is in Session 3 folder
3)  Compute the total area of each land use class within each administrative level
4)  Plot the results. Hint:use ggplot2 library
5)  Spatial plot. Hint: use the function dcast
6)  Prepare a dynamic plot the results
