---
title: "LGEO2185: R Project Structure & Workflow -- Assignment"
author: "Kristof Van Oost, Antoine Stevens & Valentin Charlier"
Date: 05/02/2026
---
:::assignment

# Assignment 

Write an R function that gives you the predicted change in precipitation and temperature for a region, but the future rainfall should represent the mean of different models. the function should be able to accept an argument for with specific model should be used. The possible CMIP6 models are one of: `"ACCESS-CM2", "ACCESS-ESM1-5", "AWI-CM-1-1-MR", "BCC-CSM2-MR", "CanESM5", "CanESM5-CanOE", "CMCC-ESM2", "CNRM-CM6-1", "CNRM-CM6-1-HR", "CNRM-ESM2-1", "EC-Earth3-Veg", "EC-Earth3-Veg-LR", "FIO-ESM-2-0", "GFDL-ESM4", "GISS-E2-1-G", "GISS-E2-1-H", "HadGEM3-GC31-LL", "INM-CM4-8", "INM-CM5-0", "IPSL-CM6A-LR", "MIROC-ES2L", "MIROC6", "MPI-ESM1-2-HR", "MPI-ESM1-2-LR", "MRI-ESM2-0", "UKESM1-0-LL"`.

You can create this function by combining the functions described above. Also create a clear graphical representation of your workflow
:::

::: {.callout-important icon="false" collapse="true"}
## Answer

```{r setup}
# Load required libraries
library(ggplot2)
library(terra)
library(geodata)
library(RColorBrewer)
library(dplyr)
library(mapview)
library(leafsync)
library(here)

# Set default paths using the 'here' package
# 'here()' ensures paths are relative to the project root, making the code more portable
#options(geodata_default_path = here::here())
#geodata_path() # Verify the geodata path

# Set geodata cache folder
options(geodata_default_path = file.path("data", "raw"))

# create folder if missing
dir.create(file.path("data", "raw"), recursive = TRUE, showWarnings = FALSE)

# check
geodata_path()
```

```{r}
# Load administrative boundary data for the Democratic Republic of Congo (DRC)
# gadm() retrieves administrative boundary data at specified resolution and level
DRC <- gadm(country = "COD", level = 0, resolution = 1)

#' Estimate Changes in Precipitation Using CMIP Model Scenarios
#'
#' @description Function to estimate changes in precipitation based on CMIP model scenarios.
#' @param ROI Region of Interest as a SpatVector.
#' @param res Resolution of the data (2.5, 5, or 10).
#' @param ssp Shared Socio-economic Pathway code ("126", "245", "370", or "585").
#' @param models Vector of CMIP6 models to use (e.g., "ACCESS-CM2", "HadGEM3-GC31-LL").
#' @param year Future year/period ("2021-2040", "2041-2060", "2061-2080").
#' @return A list containing two rasters: absolute change and relative change (in mm).
#' @examples
#' DRC <- gadm(country = "COD", level = 0, resolution = 1)
#' models <- c("ACCESS-CM2", "HadGEM3-GC31-LL")
#' precip_change <- Get_Future_ensemble_Precipitation(DRC, 5, "585", models, "2041-2060")
Get_Future_ensemble_Precipitation <- function(ROI, res, ssp, models, year) {
    # Makse sure we have the required packages within the function
    require(terra)
    require(geodata)
    require(dplyr)

    # Validate ROI
    if (!inherits(ROI, "SpatVector")) {
        stop("The 'ROI' parameter must be a SpatVector object.")
    }

    # Download current precipitation data and clip to ROI
    # worldclim_global() retrieves current climate data for the specified variable and resolution
    current <- worldclim_global(var = "prec", res = res) |>
        crop(ROI) |> # Crop to the specified region of interest
        mask(ROI) |> # Mask areas outside the ROI
        sum() # Sum across all layers (annual total precipitation)

    # Initialize a list for future precipitation data
    future <- vector("list", length(models))

    # Loop through each model to calculate future precipitation
    for (i in 1:length(models)) {
        model <- models[i]
        print(model) # Display the current model being processed
        # Retrieve future climate data for the specified model, SSP, and year, then clip and mask to ROI
        future[[i]] <- cmip6_world(model, ssp, year, var = "prec", res = res) |>
            crop(ROI) |>
            mask(ROI) |>
            sum() # Sum across all layers (annual total precipitation)
    }

    # ALTERNATIVE: This is even more simple (but adds a dependency)
    # require(purrr)
    # future <- map(models,~ cmip6_world(.x,ssp, year, var = "prec", res = res) %>%
    #                           crop(ROI) %>%
    #                           mask(ROI) %>%
    #                           sum())

    # Cast the list to a raster
    future <- rast(future)
    # Compute the mean precipitation across all models
    future <- mean(future)

    # Calculate absolute change (total precipitation difference) and relative change (percentage difference)
    delta_annual <- sum(future - current)
    delta_prec <- sum(future - current) / sum(current)

    return(list(abs_change = delta_annual, rel_change = delta_prec))
}

Get_Future_ensemble_Precipitation_FAST <- function(ROI, res, ssp, models, year) {

  require(terra)
  require(geodata)

  # Validate ROI
  if (!inherits(ROI, "SpatVector")) {
    stop("The 'ROI' parameter must be a SpatVector object.")
  }

  # Current precipitation (annual total)
  current <- worldclim_global(var = "prec", res = res) |>
    crop(ROI) |>
    mask(ROI) |>
    sum()

  # Start with an EMPTY SpatRaster (this is what makes it slow)
  future <- rast(nlyrs = length(models))

  # Loop through models and GROW the raster by adding layers
  for (model in models) {
    message(model)

    r <- cmip6_world(model, ssp, year, var = "prec", res = res) |>
      crop(ROI) |>
      mask(ROI) |>
      sum()

    # GROW the object: each add() expands and copies internally (slow)
    add(future) <- r
  }

  # Now compute ensemble mean
  future_mean <- mean(future)

  # Changes
  delta_annual <- sum(future_mean - current)
  delta_prec   <- delta_annual / sum(current)

  list(abs_change = delta_annual, rel_change = delta_prec)
}
```

```{r}
# Define models and test the functions
# Example models to be used for analysis
models <- c("ACCESS-CM2", "HadGEM3-GC31-LL","INM-CM4-8", "INM-CM5-0", "IPSL-CM6A-LR")

# Measure execution time for the standard function
tictoc::tic()
precip_change <- Get_Future_ensemble_Precipitation(res = 5, ssp = "585", models = models, year = "2041-2060", ROI = DRC)
tictoc::toc()

# Visualize the results using mapview
# Generate color palette for visualization
cols <- colorRampPalette(brewer.pal(9, "Blues"))
# Map absolute changes
m_abs <- mapview(precip_change$abs_change, col.regions = cols, na.color = "transparent")
# Map relative changes
m_rel <- mapview(precip_change$rel_change, col.regions = cols, na.color = "transparent")
# Synchronize the two maps for side-by-side comparison
leafsync::sync(m_abs, m_rel)

# now same with growing a raster by adding layers
tictoc::tic()
precip_change <- Get_Future_ensemble_Precipitation_FAST(res = 5, ssp = "585", models = models, year = "2041-2060", ROI = DRC)
tictoc::toc()
```
:::