---
title: "LGEO2185: Designing a Geoproject -- Assignment"
author: "Kristof Van Oost, Antoine Stevens & Valentin Charlier"
Date: 07/02/2026
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
pacman::p_load(tidyverse, here)
source("R/suitability_fn.R")
```

::: assignment
# Assignment

Building on what we have done there together, the challenge is now to further develop this model and to make it broader by also predicting future suitability maps using the cmip climate predictions. Make sure that your model can predict current suitability and suitability for 2030, 2050 and 2070. For example using a loop over the years: `years<-c("1970-2000", "2021-2040", "2041-2060", "2061-2080")`

Also make sure your model can handle different type of climate models and scenario's.

The output should be a list containing leaflet maps for the current situation, leaflet maps showing the change for each interval relative to the current condition, and the suitability maps for each time interval in `spatRaster` format.
:::


::: {.callout-important icon="false" collapse="true"}
## Answer

```{r, file='R/suitability_fn.R'}
```

# Example

```{r example, eval = TRUE, warning=FALSE, message=FALSE}
pacman::p_load(tidyverse, terra, leaflet, leafsync, geodata)
# Example using the fct in R/suitability_fn.R
options(geodata_default_path = "data/raw")
region <- gadm(country = "MEX", level = 0)
output <- suitability_map(
    region = region, genus = "ambystoma",
    species = "mexicanum", res = 10,
    model = "ACCESS-CM2", ssp = "126"
)
r <- output$suitability_map
plot_leaflet(r)
```

```{r change, eval = TRUE, warning=FALSE, message=FALSE}
# Plot change
change <- list(
    "Suitability change by 2030" = r[[2]] - r[[1]],
    "Suitability change by 2050" = r[[3]] - r[[1]],
    "Suitability change by 2070" = r[[4]] - r[[1]]
)
plot_leaflet(change)
```
:::