---
title: "LGEO2185: RF classification with tidymodels using GEE-derived predictors"
authors: "Kristof Van Oost & Antoine Stevens"
Date: 03/14/2025
editor: 
  markdown: 
    wrap: 72
---

## Session learning objectives

By the end of this session, students should be able to:

1.  Understand the main building blocks of ML implementation. This
    tutorial is based on [**tidymodels**](https://www.tidymodels.org/):
    `rsample`, `recipes`, `parsnip`, `workflows`, `tune`, `yardstick`.\
2.  Build a **reproducible classification pipeline** (data split →
    recipe → model → workflow → cross-validation → tuning → final fit).\
3.  Use **k-fold cross-validation** to estimate generalization
    performance and understand why it beats a single train/test split
    for model selection.\
4.  Tune Random Forest hyperparameters (`mtry`, `min_n`, `trees`) and
    select a final model using a clear metric.\
5.  Interpret the fitted model using **variable importance** and connect
    it to geography/remote sensing reasoning (what drivers make sense,
    and why).
6.  Understand potential and limitations of decision trees and Random
    Forest ML approaches

## Introduction Decision trees and Random Forest

A decision tree is a rule-based model that partitions the predictor
space into rectangular regions using simple threshold rules. In this
example, we model **landslide occurrence (`lslpts`)** using terrain
predictors from the `spDataLarge` dataset. We fit a **small tree** (only
two decision levels) so the rules remain easy to interpret. We then
visualize the tree rules as **threshold lines** in a scatterplot.

```{r}
#| message: false
#| warning: false

data("lsl", "study_mask", package = "spDataLarge")
ta <- terra::rast(system.file("raster/ta.tif", package = "spDataLarge"))

library(rpart)
library(rpart.plot)
library(ggplot2)
library(patchwork)
library(png)
library(grid)

set.seed(1)

tree2 <- rpart(
  lslpts ~ slope + cplan + cprof + elev + log10_carea,
  data = lsl,
  method = "class",
  control = rpart.control(
    maxdepth = 2,
    cp = 0.0,
    minsplit = 20
  )
)

tree2
```

```{r}
#| echo: false
# --- 1) Render tree to a PNG (base plot) ---
tree_png <- tempfile(fileext = ".png")
png(tree_png, width = 900, height = 600, res = 130)
rpart.plot(tree2, type = 2, extra = 104, fallen.leaves = TRUE, box.palette = "Blues")
dev.off()

# Read PNG and wrap as a patchwork-able element
img <- png::readPNG(tree_png)
p_tree <- patchwork::wrap_elements(full = grid::rasterGrob(img, interpolate = TRUE))

# --- 2) Scatter plot with clearer TRUE/FALSE colors ---
p_scatter <- ggplot(lsl, aes(x = slope, y = cplan, color = lslpts)) +
    geom_point(alpha = 0.6, size = 2) +
  geom_vline(xintercept = 35, 
             linetype = "dashed", 
             linewidth = 1.0, 
             color = "black") +
  
  geom_segment(aes(x = min(slope), xend = 35,
                   y = -0.095, yend = -0.095),
               inherit.aes = FALSE,
               linetype = "dashed",
               linewidth = 1.0,
               color = "black") +
  
  geom_segment(aes(x = 35, xend = max(slope),
                   y = -0.02, yend = -0.02),
               inherit.aes = FALSE,
               linetype = "dashed",
               linewidth = 1.0,
               color = "black") +
  
  scale_color_manual(values = c("FALSE" = "#1f78b4",
                                "TRUE"  = "#e31a1c")) +
  
  theme_minimal(base_size = 12)

# --- Make sure scatter has a proper legend ---
p_scatter <- p_scatter +
  labs(color = "Landslide") +           # legend title (optional)
  guides(color = guide_legend(title = NULL))  # remove title if you prefer

# --- (Optional) give both panels similar margins for better vertical alignment ---
p_scatter <- p_scatter + theme(plot.margin = margin(5, 5, 5, 5))
p_tree    <- p_tree    + theme(plot.margin = margin(5, 5, 5, 5))

# --- Two-panel figure with legend ABOVE the whole figure ---
(p_tree | p_scatter) +
  plot_layout(
    widths = c(1.8, 1),
    guides = "collect"     # collect legend(s) into one
  ) +
  plot_annotation(
    theme = theme(
      legend.position  = "top",
      legend.direction = "horizontal",
      legend.box       = "horizontal"
    )
  )
  
```

To understand *why* decision trees behave differently than linear
models, it helps to look at the **geometry of the decision boundary** in
a 2-predictor space (X1–X2).

-   **Linear models** separate classes with (at most) a *single straight
    line* in predictor space.
-   **Decision trees** split the space using *axis-aligned thresholds*
    (vertical/horizontal cuts), creating **rectangular regions**.
-   As trees grow deeper, they can approximate curved/diagonal
    boundaries, but they do so via increasingly fine “steps”.

The figure below illustrates these ideas in four panels:

1.  Linear model: one straight boundary\
2.  Same linear boundary, with illustrative tree-style axis splits
    overlaid\
3.  Tree decision regions (rectangular partitions)\
4.  A tiny tree with only two criteria (two splits)

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(patchwork)
library(tibble)

# --- 1) Make a predictor grid (toy 2D feature space) ---
grid <- expand.grid(
  x1 = seq(-2.2, 2.2, length.out = 320),
  x2 = seq(-2.2, 2.2, length.out = 320)
)

# --- 2) Define a LINEAR classifier (one diagonal boundary) ---
grid$cls_linear <- ifelse(grid$x2 >= 0.6 * grid$x1 - 0.4, "A", "B")

# --- 3) Define a TREE-like classifier (axis-aligned rectangles) ---
grid$cls_tree <- with(grid, ifelse(
  x1 < -1.0,
  ifelse(x2 >= -1.2, "A", "B"),
  ifelse(
    x1 < 0.8,
    ifelse(x2 >= -0.3, "A", "B"),
    ifelse(x2 >= 0.4, "A", "B")
  )
))

# --- 4) Define a very small 2-split tree example ---
grid$cls_tree2 <- with(grid, ifelse(x1 < -1, "A", ifelse(x2 < 1, "B", "A")))

# --- Colors (light blue vs sand) ---
fill_cols <- c(A = "#A7D0E3", B = "#F2D6A2")

# --- Shared styling: no titles ---
thm <- theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    legend.position = "none",
    plot.title = element_blank()
  )

p1 <- ggplot(grid, aes(x1, x2, fill = cls_linear)) +
  geom_raster() +
  geom_abline(slope = 0.6, intercept = -0.4, linewidth = 0.7) +
  scale_fill_manual(values = fill_cols) +
  coord_equal() +
  labs(x = "X1", y = "X2") +
  thm

p2 <- ggplot(grid, aes(x1, x2, fill = cls_linear)) +
  geom_raster(alpha = 0.95) +
  geom_abline(slope = 0.6, intercept = -0.4, linewidth = 0.6) +
  geom_vline(xintercept = c(-1.0, 0.8), linewidth = 0.6) +
  geom_hline(yintercept = c(-1.2, -0.3, 0.4), linewidth = 0.6) +
  scale_fill_manual(values = fill_cols) +
  coord_equal() +
  labs(x = "X1", y = "X2") +
  thm

p3 <- ggplot(grid, aes(x1, x2, fill = cls_tree)) +
  geom_raster() +
  geom_vline(xintercept = c(-1.0, 0.8), linewidth = 0.7) +
  geom_segment(aes(x = -2.2, xend = -1.0, y = -1.2, yend = -1.2), linewidth = 0.7) +
  geom_segment(aes(x = -1.0, xend = 0.8,  y = -0.3, yend = -0.3), linewidth = 0.7) +
  geom_segment(aes(x = 0.8,  xend = 2.2,  y = 0.4,  yend = 0.4),  linewidth = 0.7) +
  scale_fill_manual(values = fill_cols) +
  coord_equal() +
  labs(x = "X1", y = "X2") +
  thm

p4 <- ggplot(grid, aes(x1, x2, fill = cls_tree2)) +
  geom_raster() +
  geom_vline(xintercept = -1, linewidth = 0.7) +
  geom_hline(yintercept = 1, linewidth = 0.7) +
  scale_fill_manual(values = fill_cols) +
  coord_equal() +
  labs(x = "X1", y = "X2") +
  thm

(p1 | p2) / (p3 | p4)
```

Top-left: a linear decision boundary (one straight line). Top-right: the
same linear boundary, with tree-style axis-aligned split lines overlaid
(vertical/horizontal thresholds). Bottom-left: the resulting tree
decision regions (rectangular partitions). Bottom-right: a minimal tree
with only two split criteria (very interpretable, but limited
flexibility). This illustrates the key geometric difference: linear
models are global and smooth, while trees build step-wise, rectangular
approximations by repeatedly applying simple threshold rules.Note that
decision trees can also be applied to both classification and regression
problems. In the case of regression problems, the predicted value
becomes the average value of the partition.

Decision trees are unstable (high variance) and are also sensitive to
small data changes. This is illustrated below: we fit the trees across
multiple random splits. The figure shows that the Training AUROC (see
session 4) is stable, while the AUROC of the test models varies a lot.

```{r}
library(rsample)
library(pROC)
library(purrr)

set.seed(1)

splits <- replicate(30, initial_split(lsl, prop = 0.75), simplify = FALSE)

aucs <- map_dfr(splits, function(spl) {
  train_dat <- training(spl)
  test_dat  <- testing(spl)
  m <- rpart(lslpts ~ slope + cplan + cprof + elev + log10_carea,
             data = train_dat,
             method = "class")
  prob_train <- predict(m, train_dat, type = "prob")[,2]
  prob_test  <- predict(m, test_dat, type = "prob")[,2]
  tibble(
    auc_train = as.numeric(auc(roc(train_dat$lslpts, prob_train, quiet = TRUE))),
    auc_test  = as.numeric(auc(roc(test_dat$lslpts, prob_test, quiet = TRUE)))
  )
})

ggplot(aucs) +
  geom_density(aes(x = auc_train, color = "Train")) +
  geom_density(aes(x = auc_test, color = "Test")) +
  theme_minimal()
```

::: callout-note
**Advantages and Limitations of Decision Trees**

Decision trees, whether used for regression or classification, offer
several important advantages compared to more classical modelling
approaches such as linear regression or logistic regression.

**Advantages** - **Interpretability**\
Trees are extremely easy to explain. In many cases, they are even more
intuitive than linear models because they rely on simple *if–then*
rules. - **Human-like reasoning**\
The hierarchical structure of trees often resembles human
decision-making processes: we tend to evaluate one criterion, then
refine based on another. - **Graphical representation**\
Trees can be visualised directly as diagrams, making them accessible
even to non-experts, especially when the tree is small and shallow. -
**Handling of qualitative predictors**\
Trees naturally incorporate categorical variables without requiring the
creation of dummy variables or contrast coding.

**Limitations** - **Predictive performance**\
A single decision tree often does not achieve the same predictive
accuracy as more flexible or regularised statistical learning methods. -
**Instability (non-robustness)**\
Trees can be sensitive to small changes in the data. A slight
modification in the training dataset may result in a substantially
different tree structure.
:::

**From Single Trees to Tree Ensembles:** Many of these limitations can
be addressed by combining multiple trees into an ensemble. Methods such
as:

-   **Bagging**
-   **Random Forests**
-   **Boosting**

aggregate large numbers of trees to improve predictive performance and
stability. A natural way to reduce the variance and increase the test
set accuracy of a statistical learning method is to take many training
sets from the population, build a separate prediction model using each
training set and then average the resulting predictions. This is called
bagging and this is particularly useful for decision trees. This
approach has been demonstrated to substantially improve accuracy by
combining hundreds or even thousands or trees into a single prediction
model.

Bagging can improve the accuracy, but this comes at a cost as the
decision trees are no longer simple and interpretable. So the improved
accuracy comes at the expense of interpretability.

**Random Forests** provide a further improvement over bagged trees: In
each split, only a random sample of the predictors are allowed to be
used. This is counter-intuitive but this process decorrelates the trees
and results in more robust predictions.

Here we create both a logistic regression as well as an RF model for our
landslide problem using the tidymodel approach ([cfr Session
4](https://uclouvain-geog.github.io/LGEO2185/ML/Machine_Learning_in_Geosciences.html))
:

```{r}
#| message: false
#| warning: false

data("lsl", package = "spDataLarge")

library(tidymodels)
library(dplyr)
library(ggplot2)
library(patchwork)

set.seed(1)

# Ensure outcome is a factor
lsl <- lsl %>% mutate(lslpts = as.factor(lslpts))

# --- 1) Train/test split ---
lsl_split <- initial_split(lsl, prop = 0.75, strata = lslpts)
lsl_train <- training(lsl_split)
lsl_test  <- testing(lsl_split)

# --- 2) 5-fold CV on training set ---
lsl_folds <- vfold_cv(lsl_train, v = 5, strata = lslpts)

# --- 3) A common recipe (same preprocessing for both models) ---
lsl_recipe <- recipe(lslpts ~ slope + cplan + cprof + elev + log10_carea,
                     data = lsl_train) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_nzv(all_predictors())

# --- 4) Model specs ---
glm_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

rf_spec <- rand_forest(
  trees = 500,
  mtry  = tune(),     # we'll set a simple default below (no full tuning yet)
  min_n = 5
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# A simple non-tuned choice for mtry (sqrt(p) is common)
p <- 5
rf_spec <- finalize_model(rf_spec, tibble(mtry = floor(sqrt(p))))

# --- 5) Workflows ---
glm_wf <- workflow() %>%
  add_recipe(lsl_recipe) %>%
  add_model(glm_spec)

rf_wf <- workflow() %>%
  add_recipe(lsl_recipe) %>%
  add_model(rf_spec)

```

Now let's have a look at the hold-out test performance using the same
split for both models

```{r}
#| message: false
#| warning: false
mets <- metric_set(accuracy, roc_auc)
glm_last <- last_fit(glm_wf, split = lsl_split, metrics = mets)
rf_last  <- last_fit(rf_wf,  split = lsl_split, metrics = mets)

test_comp <- bind_rows(
  glm_last %>% collect_metrics() %>% mutate(model = "GLM (logistic)"),
  rf_last  %>% collect_metrics() %>% mutate(model = "Random Forest")
) %>%
  select(model, .metric, .estimate)

test_comp

```

Interpretation

Logistic regression fits a smooth, global relationship (linear in the
predictors after transformation).

Random Forest combines many trees, capturing nonlinearities and
interactions automatically.

In many geoscience susceptibility problems, RF typically achieves higher
AUROC, especially when relationships are nonlinear.

::: callout-note
Note that we did not fine-tune the (hyper)parameters of the RF model
yet. For example the nubmer of predictors sampled at each sple `mtry`,
the number of trees in the forest `trees`, and the minimum observations
per tree (`min_n`are typical parameters that can also be tuned. We will
develop this forther in a machine learning workflow below.
:::

## RF tutorial: land cover mapping

As we discussed in session 4, Tidymodels is a **coherent “grammar”** for
machine learning in R: rather than every model/package having its own
interface, tidymodels standardizes the steps you do again and again,
i.e. splitting, pre-processing, training, tuning, and evaluation.

This module follows the workflow described in [Rebecca Barter’s
tidymodels
tour](https://bookdown.org/paul/ai_ml_for_social_scientists/06_01_ml_with_tidymodels.html)
(@bauer2023tidymodels) and adapts it to **spatial training data**. We
will re-use the training data and predictors we created in Google Earth
Engine (Session Demo GEE).

```{r}
#| message: false
#| warning: false
# Core spatial + ML workflow packages
library(terra)

library(tidymodels)  # loads rsample, recipes, parsnip, yardstick, workflows, etc.
library(tune)
library(workflows)

# Optional but often useful
library(dplyr)
library(ggplot2)

# For variable importance with ranger models
library(vip)
```

To keep things reproducible we seed:

```{r}
set.seed(42)
```

------------------------------------------------------------------------

## 1. Load data (your spatial training set)

We will use a shapefile exported from GEE with training polygons/points
and predictors for a small study site near the town of Ilebo in the DR
Congo. Put the data (shapefile- and geotif-files into a subfolder
`/GEE_Exports_Annual`). The shapefile contains small polygons that were
digitized in Google Earth on high resolution satellite imagery and where
an opererator labelled the polygons according to five possible land use
classes:

```{r}
# path to the shapefile
shp_path <- "GEE_Exports_S2_Ilebo/ilebo_polygon_band_values.shp"
# load shapefile (SpatVector)
kasai_observations <- terra::vect(shp_path)
names(kasai_observations)
kasai_observations
```

A quick visualization of the data we are going to work with:

```{r, echo=FALSE}
library(leaflet)
library(sf)

kasai_sf_ll <- st_as_sf(kasai_observations) |>
st_make_valid() |>
st_transform(4326)

bb <- st_bbox(kasai_sf_ll)
center_lng <- as.numeric((bb["xmin"] + bb["xmax"]) / 2)
center_lat <- as.numeric((bb["ymin"] + bb["ymax"]) / 2)

m <- leaflet(kasai_sf_ll, options = leafletOptions(preferCanvas = TRUE)) |>
addProviderTiles(providers$Esri.WorldImagery) |>
setView(lng = center_lng, lat = center_lat, zoom = 10) |>
addPolygons(color = "yellow", weight = 2, fillOpacity = 0)

m
```

We convert to a data.frame for tidymodels as Tidymodels works on tabular
data. We typically:

-   keep geometry in a separate object
-   model on attribute table

```{r}
kasai_df <- as.data.frame(kasai_observations)
kasai_df$class <- as.factor(kasai_df$class) #for modelling later outcome should be a factor
# quick check
#glimpse(kasai_df)
```

**Note:** if you later want spatial CV, you will need
coordinates/geometry *somewhere* (even if you don’t model on them). We
keep `kasai_observations` for that purpose.

------------------------------------------------------------------------

## 2. EDA (exploration + quality checks)

### 2.1 Inspect columns + class label

The observed landcover classes in the database are given in the column
`class`. Let's explore what we have:

```{r}
# What are the unique classes?
table(kasai_df$class, useNA = "ifany")
```

### 2.2 Missing values and obvious issues

```{r}
# Count NAs per column (largest first)
na_counts <- sort(colSums(is.na(kasai_df)), decreasing = TRUE)
head(na_counts, 20)
```

It looks like our database is fully populated.

### 2.3 Quick sanity plot

```{r}
# Example: NDVI
if ("NDVI" %in% names(kasai_df)) {
  ggplot(kasai_df, aes(x = NDVI, fill = class)) +
    geom_histogram(bins = 30, alpha = 0.7) +
    theme_minimal()
}
```

What do we see here? built-up mostly under 0.5 and confirms mixed pixels
as it aggregates houses + trees and gardens. Note overlap between
cropland and savannah, this points to the need to look at multiple
predictors.

```{r}
# Example: BSI
if ("BSI" %in% names(kasai_df)) {
  ggplot(kasai_df, aes(x = BSI, fill = class)) +
    geom_histogram(bins = 30, alpha = 0.7) +
    theme_minimal()
}

```

------------------------------------------------------------------------

::: callout-note
Now we are getting to the core. The tidymodels standardizes the steps
you do again and again—splitting, pre-processing, training, tuning, and
evaluation.

A helpful overview is:

-   3 `rsample`: how you create training/test sets and resamples (folds)
-   4 `recipes`: how you define preprocessing steps
-   5`parsnip`: a consistent interface to many model engines (e.g.,
    `ranger`)
-   6 `workflows`: bundle recipe + model into a single object you can
    fit
-   7-9 `tune`: systematic tuning across resamples
-   10-11 `yardstick`: model performance metrics
:::

## 3. Split into training and testing

Use a stratified split to preserve class proportions in both sets using
`initial_split` from the `rsample` package.

```{r}
kasai_split <- initial_split(kasai_df, prop = 0.8, strata = class)

kasai_train <- training(kasai_split)
kasai_test  <- testing(kasai_split)

nrow(kasai_train); nrow(kasai_test)
```

------------------------------------------------------------------------

## 4. Recipe (preprocessing)

A recipe is **trained only on the training set** and then applied to
test/validation data.

### 4.1 Choose predictors

-   Exclude identifiers (`system:index`, `date`, etc.)
-   Exclude geometry fields if present (but we dealt with this before)
-   Keep only numeric predictors (for RF it is fine to mix factors too,
    but let's keep it simple for now)

```{r}
# drop  non-predictor fields (this is why we explored the dataset above)
drop_cols <- c("date", "yymm", "site", "year")

kasai_train2 <- kasai_train %>% select(-any_of(drop_cols))

kasai_test2 <- kasai_test %>% select(-any_of(drop_cols))

#glimpse(kasai_train2)
```

Note to self: this could have been done in step 3 just before the
initial splitting.

### 4.2 Build the recipe

> **Note:** Random Forest does not require normalized predictors in most
> cases. Still, we often: - remove near-zero variance predictors\
> - handle missing values\
> - encode categorical predictors if any

```{r}
rf_recipe <- recipe(class ~ ., data = kasai_train2) %>%
  step_zv(all_predictors()) %>%           # remove zero-variance predictors
  step_nzv(all_predictors()) %>%          # remove near-zero variance predictors
  step_impute_median(all_numeric_predictors())  # simple imputation for NA, not relevant here
```

The last part of the recipe
`step_impute_median(all_numeric_predictors())` is just for illustration
as we have no missing values. This function replaces missing values in a
dataset (NA) in numeric predictor variables by replacing them with the
median value, computed from the training data only.

------------------------------------------------------------------------

## 5. Model specification

In [tidymodels](https://www.tidymodels.org/), you specify **what kind of
model** (random forest classifier in our case) and then choose an
**engine** (`ranger`) that actually fits it.

We’ll tune three parameters (see introduction session to Machine
Learning):

-   `mtry`: number of predictors sampled at each split\
-   `min_n`: minimum node size\
-   `trees`: number of trees

```{r}
rf_model <- rand_forest(
  mode  = "classification",
  mtry  = tune(),
  min_n = tune(),
  trees = 550
) %>%
  set_engine("ranger", importance = "impurity")  # enables variable importance citeturn0search8

rf_model
```

------------------------------------------------------------------------

## 6. Workflow (recipe + model)

A workflow glues preprocessing + model together into a single object you
can resample, tune, and fit.

```{r}
rf_workflow <- workflow() %>%
  add_recipe(rf_recipe) %>%
  add_model(rf_model)

rf_workflow
```

------------------------------------------------------------------------

## 7. Cross-validation

Cross-validation gives a more stable estimate of performance than a
single split (and is the standard approach for tuning).

```{r}
set.seed(42)
kasai_folds <- vfold_cv(kasai_train2, v = 5, strata = class)

kasai_folds
```

------------------------------------------------------------------------

## 8. Tuning grid

At this step, we do not fit just one Random Forest model. Instead, we
test several versions of the model with different settings. This is
called **hyperparameter tuning**.

With the `ranger` engine, the most important parameters are usually:

-   `mtry`: the number of predictor variables randomly considered at
    each split
-   `min_n`: the minimum number of samples allowed in a terminal node
-   `trees`: the number of trees grown in the forest

The model is trained repeatedly on resampled training data, and its
predictive performance is evaluated for each parameter combination. The
goal is to identify the settings that give the best balance between
accuracy and generalization. After tuning, the best parameter
combination is selected and used to fit the final Random Forest model.We
start with a modest grid to speed up the calculations. Note that the
combination of 4 levels and 2 parameters gives 16 combinations.

::: callout-note
This is a limited analysis to speed up the calculations, under normal
circumstances you would test much more combinations (eg include the
number of trees in the fitting and increase the levels).
:::

```{r}
rf_grid <- grid_regular(
  mtry(range = c(2, 14)),   # adjust based on number of predictors
  min_n(range = c(2, 20)),
  levels = 4
)

rf_grid
```

------------------------------------------------------------------------

## 9. Tune the model (fit across folds)

We select metrics that are appropriate for classification.

-   F1 metric (`f_meas`)

> F1 score measure the harmonic mean of precision and recall. An
> introduction to classification performance metrics can be found
> [here](https://developers.google.com/machine-learning/crash-course/classification/accuracy-precision-recall).

```{r}
rf_metrics <- metric_set(f_meas)

set.seed(42)
rf_tune <- tune_grid(
  rf_workflow,
  resamples = kasai_folds,
  grid      = rf_grid,
  metrics   = rf_metrics,
  control   = control_grid(save_pred = TRUE)
)

rf_tune
```

### 9.1 Visualize tuning results

```{r}
autoplot(rf_tune)
```

### 9.2 Select the best model

```{r}
best_params <- select_best(rf_tune, metric = "f_meas")
best_params
```

------------------------------------------------------------------------

## 10. Finalize workflow + evaluate on the test set

```{r}
rf_final_workflow <- finalize_workflow(rf_workflow, best_params)

test_model_fit <- last_fit(rf_final_workflow, split = kasai_split, metrics = rf_metrics)

test_model_fit
```

### 10.1 Test-set metrics

```{r}
collect_metrics(test_model_fit)
```

### 10.2 Confusion matrix

```{r}
test_preds <- collect_predictions(test_model_fit)

test_preds %>%
  conf_mat(truth = class, estimate = .pred_class)
```

What can we learn from this confusion matrix?

------------------------------------------------------------------------

## 11. Fit final model on all data (for deployment)

Once you’re satisfied, we refit on **all labeled data** (train + test
combined) before we deploy for mapping. Think about this: what is the
difference between tuning, performance reporting and inference?

```{r}
final_model_fit <- fit(rf_final_workflow, data = bind_rows(kasai_train2, kasai_test2))

final_model_fit
```

------------------------------------------------------------------------

## 12. Variable importance (interpretation)

Variable importance is one of the simplest ways to connect “black-box”
models back to domain reasoning. This variable importance assessment is
a key interpretation step in machine learning.

```{r}
# Extract the fitted parsnip model
rf_engine_fit <- final_model_fit  %>% extract_fit_parsnip()

# Plot variable importance
vip::vip(rf_engine_fit, num_features = 20)
```

> -   Which predictors make sense for separating your classes?
>
> -   Are any “top predictors” suspicious ?
>
> -   Do the top predictors match your conceptual understanding of the
>     landscape?

------------------------------------------------------------------------

## 13. Reduced RF model

This is all fine, but the final model uses 10 bands, so when we go to
Inference, I need spatial data for all predictors it is good to check if
you do need ALL these predictors. This can be done by advanced
procedures that evaluate the performance of models with reduced number
of predictors (eg `Recursive Feature Elimination`), here we just
manually select the 5 best predictors from the variable importance
analyses above. [Here](https://stevenpawley.github.io/recipeselectors/)
is a link to a feature selection method

```{r}
  ## ---- Reduced RF model evaluated like Section 10 (using last_fit) ----

# 0) Manually choose predictors from VIP plot
top_vars <- c("B12", "B11", "BSI", "B5", "B4")
kasai_train2_reduced <- kasai_train2 %>%
  dplyr::select(class, all_of(top_vars))

# 1) Reduced recipe
rf_recipe_reduced <- recipe(class ~ ., data = kasai_train2_reduced) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors())

# 2) Reduced workflow (reuse tuned hyperparameters)
rf_workflow_reduced <- workflow() %>%
  add_recipe(rf_recipe_reduced) %>%
  add_model(rf_model) 

rf_final_workflow_reduced <- finalize_workflow(rf_workflow_reduced, best_params)

# 3) Evaluate on test set with last_fit (same approach as Section 10)
test_model_reduced_fit <- last_fit(rf_final_workflow_reduced, split = kasai_split, metrics = rf_metrics)

# 4) Confusion matrix + metrics (same as Section 10)
test_preds_reduced <- collect_predictions(test_model_reduced_fit)

test_preds_reduced %>%
  conf_mat(truth = class, estimate = .pred_class)

collect_metrics(test_model_reduced_fit)

final_reduced_model_fit <- fit(
  rf_final_workflow_reduced,
  data = bind_rows(
    kasai_train2 %>% dplyr::select(class, all_of(top_vars)),
    kasai_test2  %>% dplyr::select(class, all_of(top_vars))
  )
)

metrics_full    <- collect_metrics(test_model_fit) %>% mutate(model = "Full RF")
metrics_reduced <- collect_metrics(test_model_reduced_fit) %>% mutate(model = "Reduced RF")

model_comparison <- bind_rows(metrics_full, metrics_reduced) %>%
  select(model, .metric, .estimate) %>%
  tidyr::pivot_wider(names_from = model, values_from = .estimate)

model_comparison


```

## 14. Inference: mapping predictions with terra

Nice — inference is the fun part 😄. Since our final reduced RF is a
tidymodels workflow fit (rf_final_fit_reduced), the simplest, robust
approach with terra is:

-   load the raster stack
-   make sure layer names match your top_vars predictor names
-   convert raster values → data.frame
-   predict() with the workflow (class + probs)
-   write results back to rasters and export GeoTIFFs

Here’s an inference block:

```{r}
# 1) Load predictor raster stack (must contain layers named like top_vars)
r <- rast("GEE_Exports_S2_Ilebo/ilebo_s2_2025_int_30m.tif")
# Keep only the predictors used by the reduced model
r_red <- r[[top_vars]]

# 2) Predict land-cover classes with the tidymodels fitted workflow

r_class <- terra::predict(
r_red,
model = final_reduced_model_fit,
type = "class",
na.rm = TRUE,
filename = "GEE_Exports_S2_Ilebo/ilebo_s2_LC.tif",
overwrite = TRUE
)

cat_tab <- levels(r_class)[[1]]
lab_col <- names(cat_tab)[2]
labs <- as.character(cat_tab[[lab_col]])

lc_cols <- c(
  built_up = "#e31a1c",
  cropland = "#FFD92F",
  forest   = "#1b9e77",
  savannah = "#66a61e",
  water    = "#1f78b4"
)

cols <- unname(lc_cols[labs])
cols[is.na(cols)] <- "grey70"

plot(r_class, col = cols, main = "Predicted land cover (Reduced RF, 2025)")



```

::: callout-note
Recipes prepare data, workflows train models, tuning chooses parameters,
last_fit() evaluates once, and final fit() is what we deploy.

| **Stage**              | **Object name**             | **Package** | **What it represents**                        | **Used for**              |
|---------------|---------------|---------------|---------------|---------------|
| **Data split**         | `kasai_split`               | `rsample`   | Definition of training vs test data           | Reproducible splitting    |
|                        | `kasai_train`               | `rsample`   | Training data table                           | Model learning            |
|                        | `kasai_test`                | `rsample`   | Test data table                               | Final evaluation          |
|                        | `kasai_folds`               | `rsample`   | Cross-validation folds                        | Hyperparameter tuning     |
| **Preprocessing**      | `rf_recipe`                 | `recipes`   | Full preprocessing pipeline                   | Feature preparation       |
|                        | `rf_recipe_reduced`         | `recipes`   | Reduced-predictor pipeline                    | Feature selection         |
| **Model spec**         | `rf_model`                  | `parsnip`   | Random Forest definition + tunable parameters | Model type only (no data) |
| **Workflow**           | `rf_workflow`               | `workflows` | Recipe + model (full)                         | Tuning, fitting           |
|                        | `rf_workflow_reduced`       | `workflows` | Recipe + model (reduced)                      | Reduced modelling         |
| **Tuning**             | `rf_tune`                   | `tune`      | CV performance for many parameter sets        | Model selection           |
|                        | `best_params`               | `tune`      | Best hyperparameter combination               | Freezing complexity       |
| **Finalized workflow** | `rf_final_workflow`         | `workflows` | Tuned & locked full workflow                  | Ready for fitting         |
|                        | `rf_final_workflow_reduced` | `workflows` | Tuned & locked reduced workflow               | Ready for fitting         |
| **Test evaluation**    | `test_model_fit`            | `yardstick` | Test-set performance (full)                   | Reporting only            |
|                        | `test_model_reduced_fit`    | `yardstick` | Test-set performance (reduced)                | Model comparison          |
| **Deployment**         | `final_model_fit`           | `parsnip`   | Full model trained on all data                | Inference / mapping       |
|                        | `final_reduced_model_fit`   | `parsnip`   | Reduced model trained on all data             | Inference / mapping       |
| **Prediction output**  | `r_class`                   | `terra`     | Raster of predicted classes                   | Mapping / analysis        |
:::

------------------------------------------------------------------------

## 15. Neural networks

This session also introduces land-cover classification with neural
networks as an alternative to Random Forest.

```{r}
library(tidymodels)
library(nnet)

nn_recipe <- recipe(class ~ ., data = kasai_train2) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

nn_model <- mlp(
  mode = "classification",
  hidden_units = 10,
  penalty = 0.001,
  epochs = 200
) %>%
  set_engine("nnet", trace = FALSE)

nn_workflow <- workflow() %>%
  add_recipe(nn_recipe) %>%
  add_model(nn_model)

nn_final_workflow <- nn_workflow

nn_test_model_fit <- last_fit(nn_final_workflow, split = kasai_split, metrics = rf_metrics)
nn_test_model_fit

nn_test_preds <- collect_predictions(nn_test_model_fit)

conf_mat(nn_test_preds, truth = class, estimate = .pred_class)
collect_metrics(nn_test_model_fit)

nn_final_model_fit <- fit(nn_final_workflow, data = bind_rows(kasai_train2, kasai_test2))
nn_final_model_fit

```

Now we mirror the RF implemenation: use MLP, tune, and evaluate the
performance library(tidymodels) library(nnet)

```{r}
library(tidymodels)
library(nnet)

nn_recipe <- recipe(class ~ ., data = kasai_train2) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

nn_model <- mlp(
  mode         = "classification",
  hidden_units = tune(),
  penalty      = tune(),
  epochs       = tune()
) %>%
  set_engine("nnet", trace = FALSE)

nn_workflow <- workflow() %>%
  add_recipe(nn_recipe) %>%
  add_model(nn_model)

nn_grid <- grid_regular(
  hidden_units(range = c(5, 20)),
  penalty(range = c(-4, -2)),
  epochs(range = c(200, 500)),
  levels = 3
)

set.seed(42)
nn_tune <- tune_grid(
  nn_workflow,
  resamples = kasai_folds,
  grid      = nn_grid,
  metrics   = rf_metrics,
  control   = control_grid(save_pred = TRUE)
)

collect_metrics(nn_tune)
show_best(nn_tune, metric = "f_meas", n = 10)

best_nn_params <- select_best(nn_tune, metric = "f_meas")
best_nn_params

nn_final_workflow <- finalize_workflow(nn_workflow, best_nn_params)

test_model_nn_fit <- last_fit(
  nn_final_workflow,
  split   = kasai_split,
  metrics = rf_metrics
)

test_model_nn_fit

nn_test_preds <- collect_predictions(test_model_nn_fit)

nn_test_preds %>%
  conf_mat(truth = class, estimate = .pred_class)

collect_metrics(test_model_nn_fit)

metrics_rf <- collect_metrics(test_model_fit) %>%
  mutate(model = "Random Forest")

metrics_nn <- collect_metrics(test_model_nn_fit) %>%
  mutate(model = "Neural Network")

model_comparison_rf_nn <- bind_rows(metrics_rf, metrics_nn) %>%
  dplyr::select(model, .metric, .estimate) %>%
  tidyr::pivot_wider(
    names_from  = model,
    values_from = .estimate
  )

model_comparison_rf_nn

final_nn_model_fit <- fit(
  nn_final_workflow,
  data = bind_rows(kasai_train2, kasai_test2)
)

final_nn_model_fit

```

------------------------------------------------------------------------

## References (key resources)

-   Rebecca Barter — “Tidymodels: tidy machine learning in R.”
    citeturn1view0\
-   tidymodels.org — Random forest workflow notes (case study).
    citeturn0search2\
-   Heather Kropp — *Environmental Data Science*, Ch. 7 (Land-cover
    classification; RF + Neural Network). citeturn2view0
