---
title: "LGEO2185 -- Introduction to dplyr and the tidyverse"
author: "Kristof Van Oost, Antoine Stevens & Valentin Charlier"
Date: 2026-02-04
--- 


```{r "setup", include=FALSE}
require("knitr")
pacman::p_load(tidyverse, here)
```


## 1. Introduction

In this session, we will explore the **tidyverse** —a suite of R packages designed to make common data science tasks more intuitive and consistent. The tidyverse philosophy emphasizes human-readable code, a consistent syntax, and functions that work together, especially thanks to the **pipe orerator** (we will see later what it means). This approach can often streamline your data analysis workflow.

### 1.1 What is the tidyverse?

-   The **tidyverse** is a collection of R packages that share a common design philosophy.
-   Core packages include:
    -   **dplyr** for data manipulation,
    -   **tidyr** for reshaping or "tidying" data,
    -   **readr** for efficient data import,
    -   **purrr** for functional programming,
    -   **ggplot2** for data visualization,
    -   **tibble** for modern data frames,
    -   **stringr** for string manipulation,
    -   **lubridate** for working with dates.
-   All these packages are designed to work together, offering consistent function names, arguments, and data structures.

### 1.2 Why use dplyr?

**dplyr** provides a "grammar of data manipulation" that helps you:

1.  Write code that reflects your thinking.\
2.  Focus on what you want to do (e.g., select columns, filter rows), not on the intricacies of R's base syntax.\
3.  Create clear workflows with the **pipe operator** `|>`, which makes it easy to chain multiple transformations.

The main verbs of dplyr—`select()`, `filter()`, `mutate()`, `arrange()`, `group_by()`, and `summarise()`—cover the most common data manipulation steps. Because these functions return tibbles (enhanced data frames), you can seamlessly pass their results to other tidyverse packages like **ggplot2** for plotting or **tidyr** for reshaping.

### 1.3 The rationale behind the pipe

The **pipe operator** `|>` (sometimes called the "forward pipe") was popularized by the **magrittr** package and is now a key part of the tidyverse workflow: In short, by using the **pipe operator** `|>`, you can chain multiple transformations without repeatedly referencing your dataset:

1.  It takes the value on the left and passes it as the first argument to the function on the right.\
2.  This often improves readability by letting you "read" the code from left to right, as if you're telling R a story:\
    "Take my dataset, then do X, then do Y, then do Z, and so on..."\
3.  It reduces repetitive references to the same dataset in your code.

Together, the pipe and dplyr's verbs make it easy to write compact, readable code that directly reflects the steps of your analysis.

:::{.callout-note}
With R starting from version 4.1.0, the built-in pipe operator `|>` serve as a replacement to the former `%>%` from the **magrittr** package. 
:::

## 2. Working with Tibbles

```{r}
# If you haven't installed the tidyverse yet, run:
# install.packages("tidyverse")

library(tidyverse)
```

**Tibbles** are a modern re-imagining of data frames. They behave like data frames but are optimized for typical data science tasks. Let's load the `starwars` dataset that comes with **dplyr**.

```{r}
# Load the starwars dataset
starwars <- dplyr::starwars

# Check its class
class(starwars)

# starwars is a tibble, so printing it looks a bit different than a regular data frame
starwars
print(starwars, n = 15) # You can choose how many rows to display
```

Notice that tibbles only show a few rows and columns to keep the console tidy. They also don't automatically convert character data to factors, which can be helpful for data manipulation.

## 3. Selecting columns

A common task is to **select** only the columns you need from a dataset. Let's compare **base R** to **dplyr**.

### 3.1 Base R approaches

```{r}
# Select columns by name in base R
selection_base1 <- starwars[, c("name", "homeworld")]

# Select columns by index in base R
selection_base2 <- starwars[, c(1, 9)]
```

Both approaches work, but we often have to keep track of which index corresponds to which variable name. This can get cumbersome for large datasets.

### 3.2 dplyr approach

```{r}
# Using dplyr::select()
selection_dplyr <- starwars |>
    select(name, homeworld)

# Alternatively, you could write:
selection_dplyr2 <- select(starwars, name, homeworld)
```

The `|>` (**pipe**) operator takes the output of the left-hand side and feeds it into the first argument of the function on the right-hand side. It can help make code more readable, especially when you chain multiple operations.

### 3.3 More advanced selection

```{r}
# Select all columns EXCEPT name
starwars |> select(-name)

# Select columns containing an underscore
starwars |> select(contains("_"))

# Select columns starting with the letter "s"
starwars |> select(starts_with("s"))
```

These specialized selection helpers (`contains`, `starts_with`) are part of **tidyselect**, which comes bundled with **dplyr**.

## 4. Filtering rows

Another frequent operation is to **filter** rows based on certain conditions (e.g., we only want characters of a certain species or homeworld).

### 4.1 Base R approaches

```{r}
# We might check which rows match a condition like so:
starwars$species == "Human"

# Then use it to subset the data frame:
selection_base <- starwars[starwars$species == "Human", ]

# If you tried this without specifying the dataset columns explicitly, it wouldn't work:
# starwars[species == "Human", ] # This fails in base R unless you attach the data
```

### 4.2 dplyr approach

```{r}
# Using dplyr to filter rows:
selection_dplyr <- starwars |>
    filter(species == "Human")
```

Filtering on **multiple conditions** in base R usually involves `&` or `|` plus `which()`. In **dplyr**, you can just separate conditions with commas.

```{r}
# Base R multiple conditions:
selection_base_multiple <- starwars[which(starwars$species == "Human" &
    starwars$homeworld == "Tatooine"), ]

# dplyr multiple conditions:
selection_dplyr_multiple <- starwars |>
    filter(species == "Human", homeworld == "Tatooine")
```

Notice how the **tidyverse** version is more concise and readable.

## 5. Combining Filter and Select

Often, you need to both select certain rows and only keep certain columns. Let's see how a **traditional** approach compares with a **dplyr** approach.

### 5.1 Base R approach:

```{r}
selection_traditional <- starwars[
    starwars$species == "Human",
    c("name", "height", "birth_year")
]
```

### 5.2 dplyr approach

```{r}
selection_dplyr3 <- starwars |>
    filter(species == "Human") |>
    select(name, height, birth_year)
```

Because of the **pipe**, it's easy to see the logical flow: first filter for humans, then select the columns we need.

## 6. Grouping and summarizing

A key feature of **dplyr** is the ability to work with grouped data. You can summarize variables by categories (e.g., species).

### 6.1 Base R approach:

1.  `aggregate()`: Groups by 'species' and applies the mean function to the height and mass columns
2.  `na.rm = TRUE` ignores missing values

```{r}
mean_stats_by_species_base <- aggregate(
    cbind(height, mass) ~ species,
    data = starwars,
    FUN = function(x) mean(x, na.rm = TRUE)
)

mean_stats_by_species_base
```

### 6.2 dplyr approach

```{r}
# Summarise mean height and mass by species, ignoring NA values:
mean_stats_by_species <- starwars |>
    group_by(species) |>
    summarise_at(vars(height, mass), list(~ mean(.x, na.rm = TRUE)))

mean_stats_by_species
```

Here's what's going on:

1.  **`group_by(species)`**: Tells R to treat each species as a group.\
2.  **`summarise_at(vars(height, mass), ...)`**: Apply a function (`mean`) to the specified columns.\
3.  **`na.rm = TRUE`**: Ignores missing values (`NA`).

Alternatively, you could write:

```{r}
mean_stats_by_species_alternative <- starwars |>
    group_by(species) |>
    summarise(across(c(height, mass), ~ mean(.x, na.rm = TRUE)))
```

Both methods work similarly. `across()` is the newer approach.

## 7. A Quick glimpse at other tidyverse packages

The **tidyverse** is more than just **dplyr**. Here are a few packages you might find useful:

-   **tidyr**: Tools for reshaping data (e.g., `pivot_longer()` and `pivot_wider()` to move between "wide" and "long" formats).\
-   **readr**: Fast reading/writing of flat files (`read_csv()`, `write_csv()`, etc.).\
-   **stringr**: String manipulation functions (e.g., `str_detect()`, `str_replace()`).\
-   **purrr**: Functional programming tools for iteration (`map()`, `map_df()`).\
-   **lubridate**: Simplifies working with dates and times.

### 7.1 Example with lubridate

```{r}
# Load lubridate if not already loaded
library(lubridate)

# Some date strings in various formats
my_dates <- c("2025-02-05", "2024-12-31", "2023/01/01")

# Parse them into actual Date objects using ymd() (year-month-day)
parsed_dates <- ymd(my_dates)
parsed_dates

# Extract the components
year(parsed_dates)
month(parsed_dates, label = TRUE)
day(parsed_dates)
```

With **lubridate**, you can quickly parse dates from messy data, extract specific components (e.g., year, month, day), and perform date arithmetic.

### 7.2 Example with tidyr

While **dplyr** focuses on row/column operations, **tidyr** focuses on reshaping data between "wide" and "long" formats.
This is especially useful when preparing data for different types of analyses or visualizations: 
Some **analyses or plots** require a "long" format, where each observation is a single row. 
Other tools or methods may require a "wide" format, where repeated measurements or metrics are in separate columns.
"Tidy" structure plays well with other tidyverse packages (like **ggplot2** or **dplyr**).

Let's take a smaller slice of the `starwars` dataset to demonstrate how you might pivot it between wide and long formats. We'll use only a few columns (name, homeworld, height, and mass) to keep things manageable.

#### 7.2.1 Subset the starwars data

```{r}
# Create a smaller starwars dataset
simple_starwars <- starwars |>
    select(name, homeworld, height, mass)

# Take a look
head(simple_starwars)
```

#### 7.2.2 Base R approach (using reshape)

**Base R** has a function called `reshape()` in the `stats` package to move between wide and long formats. It can be somewhat cumbersome compared to tidyverse functions.

1.  You typically need to create an `id` column (or something similar) to uniquely identify rows.\
2.  You specify which columns are "varying" over time or measurement, and which ones are "fixed."

```{r}
# Base R reshape requires an ID variable to keep track of rows
simple_starwars$id <- seq_len(nrow(simple_starwars))
# 1) Create a smaller starwars dataset
simple_starwars <- starwars[, c("name", "homeworld", "height", "mass")]

# 3) "Wide" to "Long" using base R reshape()
base_long <- reshape(
    data      = as.data.frame(simple_starwars),
    timevar   = "measure", # The new variable that will indicate 'height' or 'mass'
    varying   = list(c("height", "mass")), # Columns to reshape
    v.names   = "value", # New column name for the measured values
    direction = "long"
)

head(base_long)
```

To go back from long to wide in **base R**, you would again call `reshape()` and specify `direction = "wide"`.

```{r}
# "Long" back to "Wide" using base R
base_wide <- reshape(
    data      = base_long,
    timevar   = "measure",
    idvar     = c("id", "name", "homeworld"),
    direction = "wide"
)

head(base_wide)
```

#### 7.2.3 tidyr approach

With **tidyr**, you can use `pivot_longer()` and `pivot_wider()` directly. This often reads more intuitively.

```{r}
# 1) pivot_longer(): wide -> long
tidy_long <- simple_starwars |>
    pivot_longer(
        cols       = c(height, mass), # which columns to stack
        names_to   = "measure", # new column with measure name
        values_to  = "value" # new column with measure value
    )

tidy_long
```

```{r}
# 2) pivot_wider(): long -> wide
tidy_wide <- tidy_long |>
    pivot_wider(
        names_from  = measure, # which column to use for new column names
        values_from = value # which column to use for new values
    )

tidy_wide
```

Notice how you don't need to manually create an ID column; **tidyr** handles that internally. The code is shorter and more descriptive of your intent: "pivot these columns, naming them this, putting their values here."

We're back in the "wide" format, but this code is more explicit about what's happening:\
- **`names_from = measure`**: The column containing old column names ("height" and "mass") is used to create new wide-format column headers.\
- **`values_from = value`**: The column containing the numeric measurements is used to populate each row.

### 7.3 Additional tidyr functions: `separate()` and `unite()`

Apart from pivoting, **tidyr** also offers functions for splitting or combining columns:

-   **`separate()`** splits one column into multiple columns.\
-   **`unite()`** does the opposite: it combines multiple columns into one.

For instance, if you had a column like `"Skywalker, Luke"` but you want separate columns for last name and first name:

```{r}
names_df <- tibble(full_name = c("Skywalker, Luke", "Solo, Han", "Organa, Leia"))

# separate()
separated_df <- names_df |>
    separate(full_name, into = c("last_name", "first_name"), sep = ", ")

separated_df
```

To recombine:

```{r}
united_df <- separated_df |>
    unite(col = "full_name_again", last_name, first_name, sep = ", ")

united_df
```

There are many other cool functions in **tidyr** for reshaping and cleaning data. Have a look at the cheatsheets [here](https://rstudio.github.io/cheatsheets/tidyr.pdf) for more!

### 7.4 Links to `ggplot2`

In `ggplot2`, data must often be in a long format to properly map variables to aesthetic attributes like `x`, `y`, `color`, or `fill`.

```{r}
starwars_weights <- starwars |>
    filter(!is.na(mass) & mass < 1000) |>
    mutate(
        weight_earth = mass, # Weight on Earth
        weight_mars = mass * 3.71 / 9.81
    ) # Weight on Mars

# Reshape data to long format
long_weights <- starwars_weights |>
    select(name, weight_earth, weight_mars) |>
    pivot_longer(
        cols = c(weight_earth, weight_mars),
        names_to = "planet",
        values_to = "weight"
    ) |>
    mutate(planet = gsub("weight_", "", planet)) # Clean up planet names

# Plot the weights on Earth and Mars
ggplot(long_weights, aes(x = name, y = weight, fill = planet)) +
    geom_bar(stat = "identity", position = "dodge") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(
        title = "Character Weights on Earth vs. Mars",
        x = "Character",
        y = "Weight (N)", # Weight in Newtons
        fill = "Planet"
    )
```

## 8. Functional programming with purrr

Another key component of the tidyverse is **purrr**, which provides a cohesive set of **functional programming** tools that make iteration in R more consistent and expressive. In particular, **purrr** excels at replacing repetitive **for loops** or `*apply` functions. In particular, it allows:

-   **Repeated operations** on each element of a list or vector.

-   **Consistent output types** (numeric, character, data frame, etc.) are desired.

-   **Parallel iteration** over multiple objects (`map2()`, `pmap()`).

-   **Robust error handling** or partial success from iterative processes (`safely()`, `possibly()`).

If your data fits neatly in rows/columns, **dplyr** verbs might be enough. But for more complex, **list-based** workflows—especially when columns contain nested lists or when you're doing more elaborate programming—**purrr** is a powerful toolkit.

### 8.1 purrr vs. traditional R loops or \*apply family

-   **Loops**: Flexible but can be verbose. You usually create an empty object, iterate, and assign results.

-   **`lapply()/sapply()`**: Shorter syntax than loops, but return type can vary (`sapply` can sometimes return a vector, sometimes a list if it can't simplify cleanly).

-   **`purrr::map_*()`**: You explicitly choose the **return type** (`map_dbl`, `map_chr`, etc.), making your code more predictable. The syntax often reads more cleanly, especially for anonymous inline functions (`~ expression`

### 8.2 The `map_` family of functions

At the core of **purrr** is the `map()` family of functions. They allow you to apply a function to each element of a list or vector and return the results in various formats:

-   `map()` returns a **list**.

-   `map_lgl()` returns a **logical** vector.

-   `map_int()` returns an **integer** vector.

-   `map_dbl()` returns a **double** (numeric) vector.

-   `map_chr()` returns a **character** vector.

-   `map_df()` (from older **purrr** versions) or `map_dfr()` / `map_dfc()` (preferred now) returns **data frames**.

Using the right function for the type of output you expect can help avoid unexpected data structures and reduce errors.

### 8.3 Example 1: Iterating over a vector

Let's calculate the square of each number in a vector.

#### 8.3.1 Base R approach

```{r, eval = FALSE}
# Traditional for loop
numbers <- c(1, 2, 3, 4, 5)
squared <- c()

for (i in numbers) {
    squared <- c(squared, i^2)
}

squared

# Using sapply
squared_sapply <- sapply(numbers, function(x) x^2)
squared_sapply
```

#### 8.3.2 purrr approach

```{r, eval = FALSE}
# Using purrr::map_dbl to return a numeric vector
squared_map <- map_dbl(numbers, ~ .x^2)
squared_map
```

### 8.4 Example 2: Applying functions to columns of a Data Frame

Suppose we want to calculate the mean of selected numeric columns (`height` and `mass`) in the `starwars` dataset.

#### 8.4.1 Base R approach

```{r, eval = FALSE}
# Traditional for loop
numeric_columns <- c("height", "mass")
means <- c()

for (col in numeric_columns) {
    means <- c(means, mean(starwars[[col]], na.rm = TRUE))
}

names(means) <- numeric_columns
means

# Using sapply
means_sapply <- sapply(starwars[c("height", "mass")], function(x) mean(x, na.rm = TRUE))
means_sapply
```

#### 8.4.2 purrr approach

```{r, eval = FALSE}
# Using purrr::map_dbl
means_map <- map_dbl(starwars[c("height", "mass")], ~ mean(.x, na.rm = TRUE))
means_map
```

### 8.5 Mapping over multiple inputs with `map2()` and `pmap()`

If your function needs **two** arguments that vary (for example, a data vector and a threshold vector), you can use **`map2()`**. For **three or more** varying arguments, **`pmap()`** is your friend. These functions let you iterate in parallel over multiple inputs. **`map2()`** takes two lists (`.x` and `.y`) in parallel. For each index, it applies the function to the corresponding elements of both lists.

Suppose we want to calculate the Body Mass Index (BMI) for each character in the starwars dataset

```{r}
# S ince operations on data frames / tibbles are vectorized, this could be achieved with a normal mutate call
starwars_bmi <- starwars |>
    mutate(
        bmi = mass / ((height / 100)^2)
    )

# But purrr helps in functional programming
# We define a BMI function
BMI <- function(mass, height) {
    mass / ((height / 100)^2)
}
# and apply this
starwars_bmi <- starwars |>
    mutate(
        bmi = map2_dbl(mass, height, ~ BMI(.x, .y))
    )
```

### 8.6 Error handling

**purrr** provides helper functions like `safely()`, `possibly()`, and `quietly()` that make it easy to handle errors or unexpected outputs within a mapping sequence. Base R typically requires more manual checks or try-catch logic for robust error handling in loops or `*apply`. From each element in `results_safe`, you can inspect `$result` or `$error`. This can be very handy for debugging or skipping problematic elements.

```{r}
# Define a safe version of the BMI function
safe_bmi <- safely(function(mass, height) {
    mass / ((height / 100)^2)
})

# Filter rows with valid mass and height (still keeping some edge cases for demonstration)
starwars_filtered <- starwars |>
    mutate(height = ifelse(name == "Darth Vader", NA, height)) |> # Introduce a NA for testing
    filter(!is.na(mass))

# Use map2 and the safe BMI function
starwars_bmi <- starwars_filtered |>
    mutate(
        bmi_safe = map2(mass, height, safe_bmi), # Apply the safe BMI function
        bmi = map_dbl(bmi_safe, ~ .x$result), # Extract the result
        bmi_error = map_chr(bmi_safe, ~ ifelse(is.null(.x$error), "None", .x$error$message)) # Extract the error
    )

# View the results with errors and calculated BMI
starwars_bmi |>
    select(name, height, mass, bmi, bmi_error)
```

::: {.callout-tip}

`map` can be used to perform within groups manipulation. Let's say we want compute the **Root Mean Squared Error (RMSE)** for a regression model that predicts a character's **Body Mass Index (BMI)** based on their **evilness**, grouped by their **`homeworld`**.

-   We use `group_by()` and `nest()` (from **tidyr**) to split the dataset into manageable subsets (one per homeworld). Each group’s data is stored in a compact format, making it easy to apply custom operations to each group.

-   The `map()` function is used to apply a **linear regression model** (`lm(bmi ~ evilness)`) to each group.

-   The output for each group includes the model and the RMSE, which measures how well the model fits the data within the group.

-   After applying the regression model to each group, the results (e.g., RMSE) are stored in a tidy format using `unnest()` for easy interpretation and visualization.

```{r}
# Create a fake column for evilness
set.seed(42)
starwars <- starwars |>
    filter(!is.na(height) & !is.na(mass) & !is.na(homeworld)) |> # Filter valid data
    mutate(
        evilness = runif(n(), 1, 10), # Random evilness between 1 and 10
        bmi = mass / ((height / 100)^2) # Calculate BMI
    )

# Group by homeworld and nest the data
# Use group_by(homeworld) to group characters by their homeworld.
# Use nest() to store each group’s data as a nested data frame within a single column.
grouped_data <- starwars |>
    group_by(homeworld) |>
    nest()

# Define a function to compute regression and calculate RMSE
compute_rmse <- function(df) {
    # Perform regression of BMI ~ evilness
    model <- lm(bmi ~ evilness, data = df)

    # Calculate RMSE
    predictions <- predict(model, df)
    rmse <- sqrt(mean((df$bmi - predictions)^2))

    # Return a tidy output with model summary and RMSE
    tibble(
        model = list(model),
        rmse = rmse
    )
}

# Apply the function to each group using map
# Use unnest() to bring the results (e.g., RMSE) back into a tidy data frame.
results <- grouped_data |>
    mutate(model_results = map(data, compute_rmse)) |>
    unnest(model_results)

# View the results
results |>
    select(homeworld, rmse)
```

:::

## 9. Conclusion

This wraps up our **introduction** to **dplyr** and the **tidyverse**!

Feel free to explore more at:\
- [dplyr.tidyverse.org](https://dplyr.tidyverse.org/)\
- [tidyverse.org](https://tidyverse.org/)

**Cheat sheets:**

-   [Data Wrangling with dplyr and tidyr](https://www.rstudio.com/wp-content%2Fuploads%2F2015%2F02%2Fdata-wrangling-cheatsheet.pdf%2F)

-   [dplyr](https://rstudio.github.io/cheatsheets/html/data-transformation.html)

-   [lubridate](https://rstudio.github.io/cheatsheets/html/lubridate.html)

-   [stringr](https://rstudio.github.io/cheatsheets/html/strings.html)

-   [tidyr](https://rstudio.github.io/cheatsheets/html/tidyr.html)

-   [purrr](https://rstudio.github.io/cheatsheets/html/purrr.html "https://rstudio.github.io/cheatsheets/html/purrr.html")

**Happy coding!**

```{r}
sessionInfo()
```