---
title: "LGEO2185: R Bindings to Geospatial Libraries --"
format: html
---

# Assignment solution

This document shows one possible solution to the assignment. The objective is to reproduce the **logic** of the Session 4 landslide-susceptibility example from `spDataLarge`, but this time by:

1. using a **STAC catalog** to discover and access the DEM,
2. using **GDAL** to assemble a working elevation raster,
3. using **QGIS/GRASS** and **WhiteboxTools** to derive the key geomorphic predictors,
4. fitting the same GLM as in the other session.

# 1. Setup

```{r}
pacman::p_load(
  sf, terra, dplyr, qgisprocess, gdalUtilities,
  whitebox, pROC, ggplot2, tidyr, rstac, tibble
)

install.packages("spDataLarge", repos = "https://geocompx.r-universe.dev")
```

```{r}
data("lsl", "study_mask", package = "spDataLarge")
ta_ref <- terra::rast(system.file("raster/ta.tif", package = "spDataLarge"))

study_mask
names(ta_ref)
```

# 2. Check the local GIS environment

```{r, include = FALSE}
# setup
Sys.setenv(
  R_QGISPROCESS_PATH = "/Applications/QGIS.app/Contents/MacOS/qgis_process",
  GRASS_PREFIX = "/Applications/GRASS-8.4.app/Contents/Resources",
  GISBASE = "/Applications/GRASS-8.4.app/Contents/Resources"
)
```

```{r}

qgisprocess::qgis_configure()
qgisprocess::qgis_providers()
whitebox::wbt_init()
```

# 3. Query a STAC catalog for a DEM

We derive the area of interest from the study mask and use it to search the Microsoft Planetary Computer STAC API for **Copernicus DEM GLO-30** tiles.

```{r}
aoi <- sf::st_bbox(study_mask)
aoi <- sf::st_transform(aoi, 4326)
aoi_vec <- c(aoi$xmin, aoi$ymin, aoi$xmax, aoi$ymax)
aoi_vec
```

```{r}
stac_items <- rstac::stac("https://planetarycomputer.microsoft.com/api/stac/v1") |>
  rstac::stac_search(
    collections = "cop-dem-glo-30",
    bbox = aoi_vec,
    limit = 50
  ) |>
  rstac::get_request()

stac_items
length(stac_items$features)
```

Planetary Computer assets usually need to be signed before access.

```{r}
stac_items_signed <- rstac::items_sign_planetary_computer(stac_items)

dem_urls <- rstac::assets_url(
  stac_items_signed,
  asset_names = "data",
  append_gdalvsi = TRUE
)

dem_urls
length(dem_urls)
```

At this stage we have a vector of remote DEM assets that GDAL can read directly.

# 4. Build a DEM mosaic with GDAL

First inspect the first source and create a VRT mosaic.

```{r}
work_dir <- file.path(tempdir(), "lgeo2185_assignment_stac")
dir.create(work_dir, recursive = TRUE, showWarnings = FALSE)

source_info <- gdalUtilities::gdalinfo(dem_urls[[1]])
source_info

vrt_file <- file.path(work_dir, "cop_dem.vrt")
dem_work <- file.path(work_dir, "cop_dem_study_area.tif")

gdalUtilities::gdalbuildvrt(
  gdalfile = dem_urls,
  output.vrt = vrt_file
)
```

Now materialize a compressed GeoTIFF from the VRT. We first print the equivalent GDAL command line.

```{r}
gdalUtilities::gdal_translate(
  src_dataset = vrt_file,
  dst_dataset = dem_work,
  projwin = c(aoi$xmin, aoi$ymax, aoi$xmax, aoi$ymin),
  of = "GTiff",
  co = c("COMPRESS=LZW"),
  dryrun = TRUE
)

gdalUtilities::gdal_translate(
  src_dataset = vrt_file,
  dst_dataset = dem_work,
  projwin = c(aoi$xmin, aoi$ymax, aoi$xmax, aoi$ymin),
  of = "GTiff",
  co = c("COMPRESS=LZW")
)
```

```{r}
dem <- terra::rast(dem_work)
plot(dem, main = "DEM assembled from Copernicus STAC assets")
plot(sf::st_geometry(study_mask), add = TRUE)
```

# 5. Derive slope and curvatures with GRASS through QGIS

```{r}
alg_id <- if (qgisprocess::qgis_has_algorithm("grass:r.slope.aspect")) {
  "grass:r.slope.aspect"
} else {
  "grass7:r.slope.aspect"
}

grass_info <- qgisprocess::qgis_run_algorithm(
  alg_id,
  elevation = dem_work,
  slope = file.path(work_dir, "slope.tif"),
  pcurvature = file.path(work_dir, "cprof.tif"),
  tcurvature = file.path(work_dir, "cplan.tif")
)

slope <- qgisprocess::qgis_as_terra(grass_info$slope)
cprof <- qgisprocess::qgis_as_terra(grass_info$pcurvature)
cplan <- qgisprocess::qgis_as_terra(grass_info$tcurvature)
elev  <- terra::rast(dem_work)

plot(slope, main = "Slope derived with GRASS")
```

`r.slope.aspect` returns several products. Here we retain the ones closest to the Session 4 predictors:

- `slope`
- profile curvature (`pcurvature` -> `cprof`)
- tangential curvature (`tcurvature` -> `cplan`)
- the DEM itself as `elev`

# 6. Derive contributing area with WhiteboxTools

We use WhiteboxTools to create a simple hydrological predictor related to contributing area.

```{r}
whitebox::wbt_init()
whitebox::wbt_wd(work_dir)

breached_file <- file.path(work_dir, "dem_breached.tif")
flowacc_file  <- file.path(work_dir, "flow_acc_cells.tif")

whitebox::wbt_breach_depressions(
  dem = dem_work,
  output = breached_file
)

whitebox::wbt_d8_flow_accumulation(
  input = breached_file,
  output = flowacc_file,
  out_type = "cells"
)

flowacc_cells <- terra::rast(flowacc_file)
cell_area <- prod(terra::res(flowacc_cells))
log10_carea <- log10(flowacc_cells * cell_area + 1)
names(log10_carea) <- "log10_carea"

plot(log10_carea, main = "log10 contributing area from WhiteboxTools")
```

This is not guaranteed to be numerically identical to the original `spDataLarge` variable, but it captures the same geomorphic idea.

# 7. Build a predictor stack and compare with the reference stack

```{r}
derived <- c(slope, cplan, cprof, elev, log10_carea)
names(derived) <- c("slope", "cplan", "cprof", "elev", "log10_carea")
derived
```

A quick visual comparison:

```{r}
par(mfrow = c(2, 2))
plot(derived[["slope"]], main = "Derived slope")
plot(ta_ref[["slope"]], main = "Reference slope")
plot(derived[["log10_carea"]], main = "Derived log10_carea")
plot(ta_ref[["log10_carea"]], main = "Reference log10_carea")
par(mfrow = c(1, 1))
```

A quick statistical comparison:

```{r}
compare_stats <- bind_rows(
  as.data.frame(terra::global(derived, c("mean", "sd"), na.rm = TRUE)) |> mutate(stack = "derived"),
  as.data.frame(terra::global(ta_ref[[names(derived)]], c("mean", "sd"), na.rm = TRUE)) |> mutate(stack = "reference")
) |>
  tibble::rownames_to_column("variable")

compare_stats
```

The values will usually differ somewhat because:

1. the STAC-sourced Copernicus DEM is not the same raster as the packaged `spDataLarge` DEM;
2. terrain derivatives vary slightly across software engines and versions;
3. hydrological preprocessing choices influence contributing-area style predictors.

# 8. Extract the predictors at the landslide sample locations

```{r}
lsl_sf <- sf::st_as_sf(lsl, coords = c("x", "y"), crs = sf::st_crs(study_mask))

extr <- terra::extract(derived, terra::vect(lsl_sf))
model_df <- bind_cols(
  lsl |> dplyr::select(lslpts),
  extr |> dplyr::select(-ID)
) |>
  tidyr::drop_na()

str(model_df)
summary(model_df)
```

# 9. Fit the Session 4 GLM

```{r}
form <- lslpts ~ slope + cplan + cprof + elev + log10_carea
fit <- glm(form, family = binomial(), data = model_df)
summary(fit)
```

Compute predicted probabilities and a simple ROC AUC:

```{r}
model_df$pred <- predict(fit, type = "response")
roc_obj <- pROC::roc(model_df$lslpts, model_df$pred, quiet = TRUE)
pROC::auc(roc_obj)
```

# 10. Optional map of susceptibility

```{r}
susc <- terra::predict(derived, fit, type = "response")
plot(susc, main = "Predicted landslide susceptibility")
plot(sf::st_geometry(study_mask), add = TRUE)
plot(sf::st_geometry(lsl_sf), add = TRUE, pch = 20, cex = 0.3)
```

# Session info

```{r}
sessionInfo()
```
